---
title: "teff_mycocosm_analysis"
format:
  html:
    toc: true
    toc-depth: 3
    theme: cosmo
    code-fold: true
    fig-align: center
  pdf:
    toc: true
    fig-pos: "H"
    geometry: margin=2cm
editor: visual
toc: TRUE
---

## Load relevant libraries

```{r}
library(phyloseq); packageVersion("phyloseq")
library(dplyr)
library(ggplot2)
library(microbiome)
library(microViz)
library(RColorBrewer)
library(gdata)
library(pairwiseAdonis)
library(ggtext)
library(kableExtra)
library(vegan)
library(paletteer)
library(colorspace)
library(emmeans)   
library(car)
library(effectsize)
library(ggVennDiagram)
library(indicspecies)
library(tidyr)
```

Loading the results from the nanopore shotgun metagenomics.

```{r}
# Read OTU and tax matrices
tax_mat <- read.table("~/Desktop/RStudio/teff_rotation_apr26/tax_matrix_0.1_kraken.txt", sep="\t", row.names=1, header=TRUE)
head(tax_mat)
otu_mat <- read.table("~/Desktop/RStudio/teff_rotation_apr26/otu_matrix_0.1_kraken.txt", sep="\t", row.names=1, header=TRUE)
head(otu_mat)
```

## Importing sample metadata

Read sample metadata BEFORE accessing otu_mat rownames

Format of sample names 1. Sample site: Dowding (D) or Trinity Hall (T) 2. Soil: soil type explained 3. Database: Standard (S) or PlusPFP (P) 4. Accuracy: SUP (S) or HAC (H)) 5. Repeat: 1 or 2 6. Confidence: 0, 0.05 or 0.1

```{r}
sample_data_df <- read.csv("~/Desktop/RStudio/teff_rotation_apr26/dna_metadata.csv", header=TRUE, row.names=1)
head(sample_data_df)
```

Now print column names of otu_mat and the row names of sample data

```{r}
colnames(otu_mat)
rownames(sample_data_df)
```

Check the sample names match the otu_mat names

```{r}
common_samples <- intersect(colnames(otu_mat), rownames(sample_data_df))
print(common_samples)

common_samples <- common_samples[common_samples != "barcode_unclassified"]
print(paste("Samples after removing unclassified:", length(common_samples)))
print(common_samples)
```

Subset common samples (only keeps samples which are common - should be everything)

```{r}
otu_mat <- otu_mat[, common_samples]
sample_data_df <- sample_data_df[common_samples, ]
```

# Removing barcodes 23 and 24

```{r}
# NOTE: These samples showed near-zero read counts across all taxa in the
# initial bar charts, indicating likely demultiplexing failure or insufficient
# DNA input. Removed prior to phyloseq construction to avoid downstream bias.

# Remove barcode23 and barcode24 by matching the full column name pattern
remove_barcodes <- c("barcode23_task23_0.1_uniq.report", 
                     "barcode24_task24_0.1_uniq.report")

# Remove from OTU matrix (columns)
otu_mat <- otu_mat[, !colnames(otu_mat) %in% remove_barcodes]

# Remove from sample metadata (rows) - adjust if your metadata uses different names
sample_data_df <- sample_data_df[!rownames(sample_data_df) %in% remove_barcodes, ]

head(sample_data_df)

# remove root 1 (redundant as we have root and introducing a number of NAs)
tax_mat <- tax_mat[, colnames(tax_mat) != "r1"]
```

## Create phyloseq objects

```{r}
OTU <- phyloseq::otu_table(as.matrix(otu_mat), taxa_are_rows = TRUE)
TAX <- phyloseq::tax_table(as.matrix(tax_mat))
colnames(TAX) <- c("Root", "Phylum", "Class", "Order",
                   "Family", "Genus", "Species", "Sub-Species")
SAMP <- phyloseq::sample_data(sample_data_df)
```

## Build phyloseq object

```{r}
ps <- phyloseq::phyloseq(OTU, TAX, SAMP)

ps
head(sample_data(ps))
sample_variables(ps)
rank_names(ps)
```

### Tidying up labels in phyloseq objects and filtering taxonomic levels

```{r}

# Keep only taxa/rows that have a genus
ps_species <- subset_taxa(ps, !is.na(Species) & Species != "")

# Making sample variables into factors 
ps_species <- ps_species %>%
  ps_mutate(
    Repeat = factor(Repeat),
    Depth = factor(Depth), 
    Genotype = factor(Genotype),
    Barcode = factor(Barcode),
    ID = factor(ID)
  )

# Keep only taxa/rows that have a genus
ps_genus <- subset_taxa(ps, !is.na(Genus) & Genus != "")
 
# Making sample variables into factors 
ps_genus <- ps_genus %>%
  ps_mutate(
    Repeat = factor(Repeat),
    Depth = factor(Depth), 
    Genotype = factor(Genotype),
    Barcode = factor(Barcode),
    ID = factor(ID)
  )

# Filtering by genotype
ps_dabbi_genus <- subset_samples(ps_genus, Genotype == "Dabbi")
ps_tsedey_genus <- subset_samples(ps_genus, Genotype == "Tsedey")

```

# Filtering out the 0.00055 least abundant reads/species

```{r}
# 2. Transform the absolute read counts into relative abundances (proportions)
# This divides each taxon's count by the total classified reads in that sample [1].
ps_rel_abun <- transform_sample_counts(ps_species, function(x) x / sum(x))

# 3. Filter the taxa based on the 0.00055 threshold.
# This keeps only taxa that have a relative abundance of at least 0.055% 
# in at least one sample.
ps_filtered <- filter_taxa(ps_rel_abun, function(x) mean(x) >= 0.00055, TRUE)

# 4. Apply this filter back to  original absolute counts 
# to work with integers for downstream analysis.
keep_taxa <- taxa_names(ps_filtered)
ps_species <- prune_taxa(keep_taxa, ps_species)

keep_taxa <- taxa_names(ps_filtered)
ps_genus <- prune_taxa(keep_taxa, ps_genus)
```

# Rarefaction for alpha diversity

```{r}
# ── Rarefaction for alpha diversity ───────────────────────────────────────
# Rarefy to a fixed, threshold of 200 reads (chosen to ensure sufficient samples pass).
# This retains all samples.
# Used ONLY for alpha diversity — beta diversity uses CLR transformation
# instead (see PCoA section).

# ── 1. Examine library sizes ──────────────────────────────────────────────
cat("=== Genus-level library sizes ===\n")
lib_genus <- sort(sample_sums(ps_genus))
print(lib_genus)
cat(sprintf("\nMin: %d  |  Median: %.0f  |  Max: %d\n",
            min(lib_genus), median(lib_genus), max(lib_genus)))

cat("\n=== Species-level library sizes ===\n")
lib_species <- sort(sample_sums(ps_species))
print(lib_species)
cat(sprintf("\nMin: %d  |  Median: %.0f  |  Max: %d\n",
            min(lib_species), median(lib_species), max(lib_species)))

# ── 2. Set fixed rarefaction threshold ───────────────────────────────────

RAREFY_DEPTH <- 200L

cat(sprintf("\nFixed rarefaction threshold: %d reads\n", RAREFY_DEPTH))
cat(sprintf("Genus-level   samples below threshold: %d\n",
            sum(lib_genus   < RAREFY_DEPTH)))
cat(sprintf("Species-level samples below threshold: %d\n",
            sum(lib_species < RAREFY_DEPTH)))

# ── 3. Remove samples below threshold ────────────────────────────────────
ps_genus_filt   <- prune_samples(sample_sums(ps_genus)   >= RAREFY_DEPTH, ps_genus)
ps_species_filt <- prune_samples(sample_sums(ps_species) >= RAREFY_DEPTH, ps_species)

n_dropped_genus   <- nsamples(ps_genus)   - nsamples(ps_genus_filt)
n_dropped_species <- nsamples(ps_species) - nsamples(ps_species_filt)

cat(sprintf("\nGenus:   %d sample(s) removed\n", n_dropped_genus))
cat(sprintf("Species: %d sample(s) removed\n",   n_dropped_species))

# ── 4. Rarefy ─────────────────────────────────────────────────────────────
set.seed(123)
ps_genus_rar <- phyloseq::rarefy_even_depth(
  ps_genus_filt,
  sample.size = RAREFY_DEPTH,
  rngseed     = 123,
  replace     = FALSE,
  trimOTUs    = TRUE,
  verbose     = TRUE
)

set.seed(123)
ps_species_rar <- phyloseq::rarefy_even_depth(
  ps_species_filt,
  sample.size = RAREFY_DEPTH,
  rngseed     = 123,
  replace     = FALSE,
  trimOTUs    = TRUE,
  verbose     = TRUE
)

cat(sprintf("\nFinal genus-level rarefied object:   %d samples, %d taxa\n",
            nsamples(ps_genus_rar), ntaxa(ps_genus_rar)))
cat(sprintf("Final species-level rarefied object: %d samples, %d taxa\n",
            nsamples(ps_species_rar), ntaxa(ps_species_rar)))

```

### Bar Chart: Species Level Comparison

```{r fig.width=10,fig.height=12, dpi=300}
# Plot bar chart (hide gap labels; keep gaps as spacing)
bar_s_myco <- ps_species %>%
  comp_barplot(
    tax_level   = "Species",
    n_taxa      = 20,
    x           = "ID",
    label       = "ID",
    palette     = distinct_palette(25, pal = "brewerPlus"),
    sample_sort = "x"
  ) +
  coord_flip() +
  theme(legend.key.size = unit(0.2, "cm"),
        legend.text     = element_text(size = 10))

plot(bar_s_myco)

# Suggestion to use tax_fix but this messed up the taxonomic rankings (some genera were put at species level and some families in genus level) so decided against using it 
```

### Bar Chart: Genus Level Comparison

```{r fig.width=10,fig.height=12, dpi=300}
# Plot bar chart (hide gap labels; keep gaps as spacing)
bar_g_myco <- ps_genus %>%
  comp_barplot(
    tax_level   = "Genus",
    n_taxa      = 20,
    x           = "ID",
    label       = "ID",
    palette     = distinct_palette(25, pal = "brewerPlus"),
    sample_sort = "x"
  ) +
  coord_flip() +
  theme(legend.key.size = unit(0.2, "cm"),
        legend.text     = element_text(size = 10))

plot(bar_g_myco)

```

### Plotting most abundant taxa without filtering

```{r fig.width=10,fig.height=12, dpi=300}
# ── 1. Get top 20 genera by overall abundance (no filtering) ──
top20_named <- names(sort(taxa_sums(ps_genus), decreasing = TRUE)[1:20])
ps_top20    <- prune_taxa(top20_named, ps_genus)

# ── 2. Prepare plot dataframe ─────────────────────────────────
plot_df <- psmelt(ps_top20) %>%
  group_by(Sample) %>%
  mutate(RelAbund = Abundance / sum(Abundance) * 100) %>%
  ungroup() %>%
  mutate(
    Depth       = factor(Depth, levels = c("Base", "Mid-section", "Tip")),
    Genotype    = factor(Genotype, levels = c("Dabbi", "Tsedey")),
    RepeatNum   = as.numeric(gsub("[A-Z]", "", ID)),
    RepeatDepth = paste(gsub("[A-Z]", "", ID), Depth)
  ) %>%
  arrange(Depth, RepeatNum) %>%
  mutate(RepeatDepth = factor(RepeatDepth, levels = unique(RepeatDepth)))

# ── 3. Taxon colour order ─────────────────────────────────────
taxon_order <- plot_df %>%
  group_by(Genus) %>%
  summarise(mean_ab = mean(RelAbund), .groups = "drop") %>%
  arrange(desc(mean_ab)) %>%
  pull(Genus)

plot_df$Genus <- factor(plot_df$Genus, levels = rev(taxon_order))
n_taxa <- nlevels(plot_df$Genus)

tax_cols <- setNames(
  colorRampPalette(brewer.pal(12, "Paired"))(n_taxa),
  levels(plot_df$Genus)
)

# ── 4. Shared theme ───────────────────────────────────────────
bar_theme <- theme_bw(base_size = 11) +
  theme(
    axis.text.x      = element_text(angle = 45, hjust = 1, size = 8),
    axis.title.x     = element_blank(),
    axis.title.y     = element_text(size = 9),
    strip.background = element_rect(fill = "grey92"),
    strip.text       = element_text(face = "bold", size = 11),
    legend.text      = element_text(face = "italic", size = 8),
    legend.key.size  = unit(0.4, "cm"),
    panel.spacing    = unit(0.8, "lines")
  )

# ── 5. Plot function per genotype ─────────────────────────────
make_panel <- function(geno, show_legend = FALSE) {
  df <- filter(plot_df, Genotype == geno)

  p <- ggplot(df, aes(x = RepeatDepth, y = RelAbund, fill = Genus)) +
    geom_col(width = 0.85, colour = NA) +
    facet_wrap(~ Depth, scales = "free_x", nrow = 1) +
    scale_fill_manual(values = tax_cols) +
    scale_y_continuous(
      expand = c(0, 0),
      limits = c(0, 101),
      breaks = c(0, 25, 50, 75, 100)
    ) +
    labs(
      y    = "Relative abundance (% of top 20 genera)",  # ✔ fixed \\n → \n
      fill = "Genus",
      title = geno
    ) +
    bar_theme +
    theme(
      plot.title = element_text(face = "bold", hjust = 1, size = 11,
                                margin = margin(b = 2))
    )

  if (!show_legend) p <- p + theme(legend.position = "none")
  p
}

# ── 6. Build and combine ──────────────────────────────────────
library(patchwork)  # ✔ required for / and & operators

p_dabbi  <- make_panel("Dabbi",  show_legend = FALSE)
p_tsedey <- make_panel("Tsedey", show_legend = FALSE)

p_combined <- (p_dabbi / p_tsedey) +
  plot_layout(guides = "collect") &  # ✔ collects legends into one
  theme(
    legend.position = "right",
    plot.margin     = margin(t = 5, r = 5, b = 5, l = 5)
  )

print(p_combined)

```

### Plotting most abundant taxa with filtering

```{r fig.width=10,fig.height=12, dpi=300}
library(patchwork)  # ✔ required for / and & operators

# ── 1. Prepare data ───────────────────────────────────────────
# GTDB unnamed lineage pattern
gtdb_unnamed <- paste0(
  "^[A-Z]{2,}[0-9]+$",
  "|^UBA", "|^TH[0-9]",
  "|^BOG-", "|^PALSA-",
  "|^FEB-", "|^DP-",
  "|^ZC4", "|^JALH", "|^JADJ",
  "|^JAJK", "|^JAEU",
  "|^CAMLFJ", "|^WHUA",
  "|^IMCC", "|^AWTP",
  "|^TA-", "|^Ch[0-9]",
  "|^Nitrospira_[A-Z]",
  "|^Ga[0-9]",
  "|^Palsa-"
)

# Remove taxa with no resolved genus name or GTDB placeholder codes
ps_named <- subset_taxa(ps_genus,
  !is.na(Genus) &
  Genus != "" &
  !grepl(gtdb_unnamed, Genus) &
  !grepl("^[Uu]nclassified|^[Uu]nknown|^uncultured", Genus)
)

# Get top 20 named taxa by overall abundance
top20_named <- names(sort(taxa_sums(ps_named), decreasing = TRUE)[1:20])

# Prune to only those 20 taxa (raw counts — no transformation yet)
ps_top20 <- prune_taxa(top20_named, ps_named)

plot_df <- psmelt(ps_top20) %>%
  group_by(Sample) %>%
  mutate(RelAbund = Abundance / sum(Abundance) * 100) %>%
  ungroup() %>%
  mutate(
    Depth       = factor(Depth, levels = c("Base", "Mid-section", "Tip")),
    Genotype    = factor(Genotype, levels = c("Dabbi", "Tsedey")),
    RepeatDepth = paste0(gsub("[A-Z]", "", ID), "", Depth)  
  )

# Fix x-axis order: sort by repeat number within each depth
plot_df <- plot_df %>%
  mutate(RepeatNum = as.numeric(gsub("[A-Z]", "", ID))) %>%
  arrange(Depth, RepeatNum) %>%
  mutate(RepeatDepth = factor(RepeatDepth, levels = unique(RepeatDepth)))

# ── 2. Consistent taxon colour order ─────────────────────────
taxon_order <- plot_df %>%
  group_by(Genus) %>%
  summarise(mean_ab = mean(RelAbund), .groups = "drop") %>%
  arrange(desc(mean_ab)) %>%
  pull(Genus)

plot_df$Genus <- factor(plot_df$Genus, levels = rev(taxon_order))

n_taxa   <- nlevels(plot_df$Genus) 
tax_cols <- setNames(
  colorRampPalette(brewer.pal(12, "Paired"))(n_taxa),
  levels(plot_df$Genus)
)

# ── 3. Shared theme ───────────────────────────────────────────
bar_theme <- theme_bw(base_size = 11) +
  theme(
    axis.text.x      = element_text(angle = 45, hjust = 1, size = 8),
    axis.title.x     = element_blank(),
    axis.title.y     = element_text(size = 9),
    strip.background = element_rect(fill = "grey92"),
    strip.text       = element_text(face = "bold", size = 11),
    legend.text      = element_text(face = "italic", size = 8),
    legend.key.size  = unit(0.4, "cm"),
    panel.spacing    = unit(0.8, "lines")
  )

# ── 4. Plot function per genotype ─────────────────────────────
make_panel <- function(geno, show_legend = FALSE) {
  df <- filter(plot_df, Genotype == geno)

  p <- ggplot(df, aes(x = RepeatDepth, y = RelAbund, fill = Genus)) +
    geom_col(width = 0.85, colour = NA) +
    facet_wrap(~ Depth, scales = "free_x", nrow = 1) +
    scale_fill_manual(values = tax_cols) +
    scale_y_continuous(
      expand = c(0, 0),
      limits = c(0, 101),
      breaks = c(0, 25, 50, 75, 100)
    ) +
    labs(
      y     = "Relative abundance (% of top 20 named genera)",
      fill  = "Genus",
      title = geno
    ) +
    bar_theme +
    theme(
      plot.title = element_text(face = "bold", hjust = 1, size = 11,
                                margin = margin(b = 2))
    )

  if (!show_legend) p <- p + theme(legend.position = "none")
  p
}

# ── 5. Build individual panels ────────────────────────────────
p_dabbi  <- make_panel("Dabbi",  show_legend = FALSE)
p_tsedey <- make_panel("Tsedey", show_legend = FALSE)

# ── 6. Combine with patchwork ─────────────────────────────────
p_combined <- (p_dabbi / p_tsedey) +
  plot_layout(guides = "collect") &  # ✔ merges legends into one
  theme(
    legend.position = "right",
    plot.margin     = margin(t = 5, r = 5, b = 5, l = 5)
  )

print(p_combined)

```

## changing panel composition to look at depths next to each other

```{r fig.width=12,fig.height=24, dpi=300}
library(patchwork)  # ✔ required for / and & operators

# ── 1. Prepare data ───────────────────────────────────────────
# GTDB unnamed lineage pattern
gtdb_unnamed <- paste0(
  "^[A-Z]{2,}[0-9]+$",
  "|^UBA", "|^TH[0-9]",
  "|^BOG-", "|^PALSA-",
  "|^FEB-", "|^DP-",
  "|^ZC4", "|^JALH", "|^JADJ",
  "|^JAJK", "|^JAEU",
  "|^CAMLFJ", "|^WHUA",
  "|^IMCC", "|^AWTP",
  "|^TA-", "|^Ch[0-9]",
  "|^Nitrospira_[A-Z]",
  "|^Ga[0-9]",
  "|^Palsa-"
)

# Remove taxa with no resolved genus name or GTDB placeholder codes
ps_named <- subset_taxa(ps_genus,
  !is.na(Genus) &
  Genus != "" &
  !grepl(gtdb_unnamed, Genus) &
  !grepl("^[Uu]nclassified|^[Uu]nknown|^uncultured", Genus)
)

# Get top 20 named taxa by overall abundance
top20_named <- names(sort(taxa_sums(ps_named), decreasing = TRUE)[1:20])

# Prune to only those 20 taxa (raw counts — no transformation yet)
ps_top20 <- prune_taxa(top20_named, ps_named)

plot_df <- psmelt(ps_top20) %>%
  group_by(Sample) %>%
  mutate(RelAbund = Abundance / sum(Abundance) * 100) %>%
  ungroup() %>%
  mutate(
    Depth       = factor(Depth, levels = c("Base", "Mid-section", "Tip")),
    Genotype    = factor(Genotype, levels = c("Dabbi", "Tsedey")),
    RepeatDepth = paste0(gsub("[A-Z]", "", ID), "", Depth)  # ✔ fixed \\\\n → \\n
  )

# Fix x-axis order: sort by repeat number within each depth
plot_df <- plot_df %>%
  mutate(RepeatNum = as.numeric(gsub("[A-Z]", "", ID))) %>%
  arrange(Depth, RepeatNum) %>%
  mutate(RepeatDepth = factor(RepeatDepth, levels = unique(RepeatDepth)))

# ── 2. Consistent taxon colour order ─────────────────────────
taxon_order <- plot_df %>%
  group_by(Genus) %>%
  summarise(mean_ab = mean(RelAbund), .groups = "drop") %>%
  arrange(desc(mean_ab)) %>%
  pull(Genus)

plot_df$Genus <- factor(plot_df$Genus, levels = rev(taxon_order))

n_taxa   <- nlevels(plot_df$Genus)  # ✔ nlevels() is safer than length(unique()) after factoring
tax_cols <- setNames(
  colorRampPalette(brewer.pal(12, "Paired"))(n_taxa),
  levels(plot_df$Genus)
)

# ── 3. Shared theme ───────────────────────────────────────────
bar_theme <- theme_bw(base_size = 11) +
  theme(
    axis.text.x      = element_text(angle = 45, hjust = 1, size = 8),
    axis.title.x     = element_blank(),
    axis.title.y     = element_text(size = 9),
    strip.background = element_rect(fill = "grey92"),
    strip.text       = element_text(face = "bold", size = 11),
    legend.text      = element_text(face = "italic", size = 8),
    legend.key.size  = unit(0.4, "cm"),
    panel.spacing    = unit(0.8, "lines")
  )

# ── 4. Plot function per depth ────────────────────────────────
make_panel <- function(depth, show_legend = FALSE) {
  df <- filter(plot_df, Depth == depth)

  p <- ggplot(df, aes(x = RepeatDepth, y = RelAbund, fill = Genus)) +
    geom_col(width = 0.85, colour = NA) +
    facet_wrap(~ Genotype, scales = "free_x", nrow = 1) +
    scale_fill_manual(values = tax_cols) +
    scale_y_continuous(
      expand = c(0, 0),
      limits = c(0, 101),
      breaks = c(0, 25, 50, 75, 100)
    ) +
    labs(
      y     = "Relative abundance (% of top 20 named genera)",  # ✔ fixed \\\\n → \\n
      fill  = "Genus",
      title = depth
    ) +
    bar_theme +
    theme(
      plot.title = element_text(face = "bold", hjust = 1, size = 11,
                                margin = margin(b = 2))
    )

  if (!show_legend) p <- p + theme(legend.position = "none")
  p
}

# ── 5. Build individual panels ────────────────────────────────
p_base  <- make_panel("Base",  show_legend = FALSE)
p_mid   <- make_panel("Mid-section", show_legend = FALSE)
p_tip   <- make_panel("Tip", show_legend = FALSE)

# ── 6. Combine with patchwork ─────────────────────────────────
p_combined <- (p_base / p_mid / p_tip) +
  plot_layout(guides = "collect") &  # ✔ merges legends into one
  theme(
    legend.position = "right",
    plot.margin     = margin(t = 5, r = 5, b = 5, l = 5)
  )

print(p_combined)
```

## Unfiltered by depth and genotype

```{r fig.width=14,fig.height=24, dpi=300}
library(patchwork)  # ✔ required for / and & operators

# ── 1. Prepare data ───────────────────────────────────────────
# Get top 20 taxa by overall abundance from ps_genus (no name filtering)
top20_all <- names(sort(taxa_sums(ps_genus), decreasing = TRUE)[1:20])

# Prune to only those 20 taxa (raw counts — no transformation yet)
ps_top20 <- prune_taxa(top20_all, ps_genus)

plot_df <- psmelt(ps_top20) %>%
  group_by(Sample) %>%
  mutate(RelAbund = Abundance / sum(Abundance) * 100) %>%
  ungroup() %>%
  mutate(
    Depth       = factor(Depth, levels = c("Base", "Mid-section", "Tip")),
    Genotype    = factor(Genotype, levels = c("Dabbi", "Tsedey")),
    RepeatDepth = paste0(gsub("[A-Z]", "", ID), " \n", Depth)  # space before newline
  )

# Fix x-axis order: sort by repeat number within each depth
plot_df <- plot_df %>%
  mutate(RepeatNum = as.numeric(gsub("[A-Z]", "", ID))) %>%
  arrange(Depth, RepeatNum) %>%
  mutate(RepeatDepth = factor(RepeatDepth, levels = unique(RepeatDepth)))

# ── 2. Consistent taxon colour order ─────────────────────────
taxon_order <- plot_df %>%
  group_by(Genus) %>%
  summarise(mean_ab = mean(RelAbund), .groups = "drop") %>%
  arrange(desc(mean_ab)) %>%
  pull(Genus)

plot_df$Genus <- factor(plot_df$Genus, levels = rev(taxon_order))

n_taxa   <- nlevels(plot_df$Genus)
tax_cols <- setNames(
  colorRampPalette(brewer.pal(12, "Paired"))(n_taxa),
  levels(plot_df$Genus)
)

# ── 3. Shared theme ───────────────────────────────────────────
bar_theme <- theme_bw(base_size = 14) +
  theme(
    axis.text.x      = element_text(angle = 45, hjust = 1, size = 14),
    axis.title.x     = element_blank(),
    axis.title.y     = element_text(size = 17),
    strip.background = element_rect(fill = "grey92"),
    strip.text       = element_text(face = "bold", size = 16),
    legend.text      = element_text(face = "italic", size = 14),
    legend.key.size  = unit(0.4, "cm"),
    panel.spacing    = unit(0.8, "lines")
  )

# ── 4. Plot function per depth ────────────────────────────────
make_panel <- function(depth, show_legend = FALSE) {
  df <- filter(plot_df, Depth == depth)

  p <- ggplot(df, aes(x = RepeatDepth, y = RelAbund, fill = Genus)) +
    geom_col(width = 0.85, colour = NA) +
    facet_wrap(~ Genotype, scales = "free_x", nrow = 1) +
    scale_fill_manual(values = tax_cols) +
    scale_y_continuous(
      expand = c(0, 0),
      limits = c(0, 101),
      breaks = c(0, 25, 50, 75, 100)
    ) +
    labs(
      y     = "Relative abundance\n(% of top 20 genera)",
      fill  = "Genus",
      title = depth
    ) +
    bar_theme +
    theme(
      plot.title = element_text(face = "bold", hjust = 1, size = 14,
                                margin = margin(b = 2))
    )

  if (!show_legend) p <- p + theme(legend.position = "none")
  p
}

# ── 5. Build individual panels ────────────────────────────────
p_base  <- make_panel("Base",  show_legend = FALSE)
p_mid   <- make_panel("Mid-section", show_legend = FALSE)
p_tip   <- make_panel("Tip", show_legend = FALSE)

# ── 6. Combine with patchwork ─────────────────────────────────
p_combined <- (p_base / p_mid / p_tip) +
  plot_layout(guides = "collect") &
  theme(
    legend.position = "right",
    plot.margin     = margin(t = 5, r = 5, b = 5, l = 5)
  )

print(p_combined)
```

### Genus Level PCoA — Bray-Curtis and Aitchison (both retained for comparison)

```{r fig.width=12, fig.height=10, dpi=300}
# ── Bray-Curtis (retained for comparison with literature) ────
pcoa_res_bc <- ordinate(ps_genus, method = "PCoA", distance = "bray")

var_exp    <- pcoa_res_bc$values$Relative_eig * 100
xlab_text <- paste0("PCoA1 (", round(var_exp[1], 1), "%)")
ylab_text <- paste0("PCoA2 (", round(var_exp[2], 1), "%)")
grp_levels <- levels(sample_data(ps_genus)$Depth)
pal        <- qualitative_hcl(length(grp_levels), palette = "Dark 3")

ord_genus_bc <- plot_ordination(ps_genus, pcoa_res_bc,
                                color = "Depth", shape = "Genotype") +
  geom_point(size = 5, alpha = 0.7) +
  stat_ellipse(aes(color = Depth, group = Depth),
               type = "t", level = 0.95, linewidth = 0.8,
               linetype = "dashed", show.legend = FALSE) +
  scale_color_manual(values = setNames(pal, grp_levels)) +
  labs(x = xlab_text, y = ylab_text,
       title = "Bray–Curtis PCoA — Depth (colour) & Genotype (shape)") +
  theme_minimal(base_size = 16) +
  theme(legend.position = "bottom")
print(ord_genus_bc)

# ── Aitchison (CLR–Euclidean) — recommended for compositional data ───────
ps_genus_clr <- microbiome::transform(ps_genus, transform = "clr")
pcoa_res_ait <- ordinate(ps_genus_clr, method = "PCoA", distance = "euclidean")

var_exp_a  <- pcoa_res_ait$values$Relative_eig * 100
xlab_ait   <- paste0("PCoA1 (", round(var_exp_a[1], 1), "%)")
ylab_ait   <- paste0("PCoA2 (", round(var_exp_a[2], 1), "%)")

ord_genus_ait <- plot_ordination(ps_genus_clr, pcoa_res_ait,
                                 color = "Depth", shape = "Genotype") +
  geom_point(size = 5, alpha = 0.7) +
  stat_ellipse(aes(color = Depth, group = Depth),
               type = "t", level = 0.95, linewidth = 0.8,
               linetype = "dashed", show.legend = FALSE) +
  scale_color_manual(values = setNames(pal, grp_levels)) +
  labs(x = xlab_ait, y = ylab_ait,
       title = "Aitchison (CLR–Euclidean) PCoA — Depth (colour) & Genotype (shape)") +
  theme_minimal(base_size = 16) +
  theme(legend.position = "bottom")
print(ord_genus_ait)
```

## PCoA plot by genotype and depth
```{r fig.width=12, fig.height=10, dpi=300}
library(ggplot2)
library(dplyr)
library(tibble)
library(phyloseq)
library(microbiome)

# CLR-transform then compute Euclidean (= Aitchison) distance
ps_genus_clr <- microbiome::transform(ps_genus, transform = "clr")
pcoa_res     <- ordinate(ps_genus_clr, method = "PCoA", distance = "euclidean")

# Create genotype_depth grouping
sample_data(ps_genus_clr)$genotype_depth <- paste(
  sample_data(ps_genus_clr)$Genotype,
  sample_data(ps_genus_clr)$Depth,
  sep = "_"
)

# Nature-style palette
colmap <- c(
  "Dabbi_Base"         = "#9ECAE1",
  "Dabbi_Mid-section"  = "#3182BD",
  "Dabbi_Tip"          = "#08519C",
  "Tsedey_Base"        = "#FDAE6B",
  "Tsedey_Mid-section" = "#E6550D",
  "Tsedey_Tip"         = "#A63603"
)

# Same shape within genotype, mapped at genotype_depth level for merged legend
shapemap <- c(
  "Dabbi_Base"         = 16,
  "Dabbi_Mid-section"  = 16,
  "Dabbi_Tip"          = 16,
  "Tsedey_Base"        = 17,
  "Tsedey_Mid-section" = 17,
  "Tsedey_Tip"         = 17
)

var_exp   <- pcoa_res$values$Relative_eig * 100
xlab_text <- paste0("PCoA1 (", round(var_exp[1], 1), "%)")
ylab_text <- paste0("PCoA2 (", round(var_exp[2], 1), "%)")

meta_df <- as(sample_data(ps_genus_clr), "data.frame") %>%
  tibble::rownames_to_column("samples") %>%
  dplyr::select(samples, genotype_depth, Genotype, Depth)

positions <- as_tibble(pcoa_res$vectors, rownames = "samples") %>%
  dplyr::rename(PCoA1 = Axis.1, PCoA2 = Axis.2) %>%
  left_join(meta_df, by = "samples")

# Function to generate a covariance ellipse manually
make_ellipse <- function(df, level = 0.68, npoints = 200) {
  if (nrow(df) < 3) return(NULL)

  center <- c(mean(df$PCoA1), mean(df$PCoA2))
  covmat <- stats::cov(cbind(df$PCoA1, df$PCoA2))

  if (any(is.na(covmat)) || det(covmat) <= 0) return(NULL)

  angles <- seq(0, 2 * pi, length.out = npoints)
  circle <- cbind(cos(angles), sin(angles))

  radius <- sqrt(qchisq(level, df = 2))
  ell <- t(center + radius * t(circle %*% chol(covmat)))

  out <- as.data.frame(ell)
  colnames(out) <- c("PCoA1", "PCoA2")
  out$genotype_depth <- unique(df$genotype_depth)
  out
}

ellipse_df <- positions %>%
  group_by(genotype_depth) %>%
  group_split() %>%
  lapply(make_ellipse, level = 0.68) %>%
  bind_rows()

pcoa_plot <- ggplot(positions, aes(x = PCoA1, y = PCoA2)) +
  geom_polygon(
    data = ellipse_df,
    aes(fill = genotype_depth, group = genotype_depth),
    alpha = 0.15,
    colour = NA,
    show.legend = FALSE
  ) +
  geom_path(
    data = ellipse_df,
    aes(colour = genotype_depth, group = genotype_depth),
    linewidth = 0.7,
    lineend = "round",
    show.legend = FALSE
  ) +
  geom_point(
    aes(colour = genotype_depth, shape = genotype_depth),
    size = 3.2,
    stroke = 0.7
  ) +
  labs(
    x = xlab_text,
    y = ylab_text,
    colour = "Genotype × Depth",
    shape  = "Genotype × Depth"
  ) +
  scale_colour_manual(values = colmap) +
  scale_fill_manual(values = colmap) +
  scale_shape_manual(values = shapemap) +
  scale_x_continuous(expand = expansion(mult = 0.12)) +
  scale_y_continuous(expand = expansion(mult = 0.12)) +
  theme_classic(base_size = 12) +
  theme(
    axis.text         = element_text(size = 11, colour = "black"),
    axis.title        = element_text(size = 12, face = "plain"),
    axis.line         = element_line(colour = "black", linewidth = 0.5),
    axis.ticks        = element_line(colour = "black", linewidth = 0.4),
    legend.position   = "right",
    legend.title      = element_text(size = 10, face = "bold"),
    legend.text       = element_text(size = 9),
    legend.key.size   = unit(0.5, "cm"),
    legend.background = element_blank(),
    plot.margin       = margin(10, 15, 10, 10)
  )

print(pcoa_plot)
```

## Identify taxa which are constant across samples (could indicate genera which are important for soil health/functioning) - core genera

```{r}
# ── 1. Core at genus level ────────────────────────────────────
otu_mat_core <- as(phyloseq::otu_table(ps_genus), "matrix")
if (!phyloseq::taxa_are_rows(ps_genus)) otu_mat_core <- t(otu_mat_core)

n_samples <- phyloseq::nsamples(ps_genus)
present_in <- rowSums(otu_mat_core > 0)

# Present in 100% of samples
core_100 <- names(present_in[present_in == n_samples])
core_genera_100 <- unique(phyloseq::tax_table(ps_genus)[core_100, "Genus"])
cat("Genera in 100% of samples:", length(core_genera_100), "\n")
print(core_genera_100)

# Present in ≥75% of samples (more relaxed threshold)
core_75 <- names(present_in[present_in >= ceiling(0.75 * n_samples)])
core_genera_75 <- unique(phyloseq::tax_table(ps_genus)[core_75, "Genus"])
cat("\nGenera in ≥75% of samples:", length(core_genera_75), "\n")
print(core_genera_75)

# ── 2. Core shared between both genotypes ─────────────────────
dabbi_present  <- rowSums(as(phyloseq::otu_table(
  phyloseq::subset_samples(ps_genus, Genotype == "Dabbi")),  "matrix") > 0)
tsedey_present <- rowSums(as(phyloseq::otu_table(
  phyloseq::subset_samples(ps_genus, Genotype == "Tsedey")), "matrix") > 0)

# Taxa present in at least one sample of EACH genotype
shared_taxa <- intersect(names(dabbi_present[dabbi_present > 0]),
                          names(tsedey_present[tsedey_present > 0]))
shared_genera <- unique(phyloseq::tax_table(ps_genus)[shared_taxa, "Genus"])
cat("Genera shared across both genotypes:", length(shared_genera), "\n")
print(shared_genera)

```

# Core species

```{r}
# ── 3. Core at species level ──────────────────────────────────
otu_mat_sp <- as(phyloseq::otu_table(ps_species), "matrix")
if (!phyloseq::taxa_are_rows(ps_species)) otu_mat_sp <- t(otu_mat_sp)

n_samples_sp <- phyloseq::nsamples(ps_species)
present_in_sp <- rowSums(otu_mat_sp > 0)

# Present in 100% of samples
core_100_sp <- names(present_in_sp[present_in_sp == n_samples_sp])
core_species_100 <- unique(phyloseq::tax_table(ps_species)[core_100_sp, "Species"])
cat("Species in 100% of samples:", length(core_species_100), "\n")
print(core_species_100)

# Present in ≥75% of samples
core_75_sp <- names(present_in_sp[present_in_sp >= ceiling(0.75 * n_samples_sp)])
core_species_75 <- unique(phyloseq::tax_table(ps_species)[core_75_sp, "Species"])
cat("\nSpecies in ≥75% of samples:", length(core_species_75), "\n")
print(core_species_75)

# ── 4. Core species shared between both genotypes ─────────────
dabbi_sp  <- rowSums(as(phyloseq::otu_table(
  phyloseq::subset_samples(ps_species, Genotype == "Dabbi")),  "matrix") > 0)
tsedey_sp <- rowSums(as(phyloseq::otu_table(
  phyloseq::subset_samples(ps_species, Genotype == "Tsedey")), "matrix") > 0)

shared_sp_taxa <- intersect(names(dabbi_sp[dabbi_sp > 0]),
                             names(tsedey_sp[tsedey_sp > 0]))
shared_species <- unique(phyloseq::tax_table(ps_species)[shared_sp_taxa, "Species"])
cat("\nSpecies shared across both genotypes:", length(shared_species), "\n")
print(shared_species)

# ── 5. Cross-reference: core species whose genus is also core ──
# These are the highest-confidence candidates for follow-up
core_tax <- as.data.frame(phyloseq::tax_table(ps_species)[core_100_sp, ])
cross_ref <- core_tax %>%
  dplyr::filter(Genus %in% core_genera_100) %>%
  dplyr::select(Genus, Species) %>%
  dplyr::distinct() %>%
  dplyr::arrange(Genus)

cat("\nSpecies that are core AND whose genus is also core:\n")
print(cross_ref)

```

## Finding the number and IDs of genera present in all samples

```{r}
# Extract the taxonomy table and make sure genus names are clean
tax <- as.data.frame(tax_table(ps_genus))
tax_table(ps_genus)[, "Genus"] <- tax$Genus

# Get the abundance (count) table
mat <- as(otu_table(ps_genus), "matrix")
if (!taxa_are_rows(ps_genus)) mat <- t(mat)

# Count in how many samples each genus appears (non-zero abundance)
present_in_samples <- rowSums(mat > 0)

# Select only the genera that appear in all samples
core_taxa <- names(present_in_samples[present_in_samples == nsamples(ps_genus)])

# To see how many present in 7 taxa, would edit to `names(present_in_samples[present_in_samples == 7])`

# Extract their genus names
core_genera <- unique(as.character(tax_table(ps_genus)[core_taxa, "Genus"]))

# View the list of core genera
print(core_genera)
length(core_genera)  # how many there are

```


## Venn diagram of shared species across genotypes

```{r fig.width=12, fig.height=10, dpi=300}
# Install if needed: install.packages("ggVennDiagram")
library(ggVennDiagram)

# 1. Extract Species lists for each Location
dabbi_taxa <- taxa_names(subset_samples(ps_species, Genotype == "Dabbi") %>% prune_taxa(taxa_sums(.) > 0, .))
tsedey_taxa <- taxa_names(subset_samples(ps_species, Genotype == "Tsedey") %>% prune_taxa(taxa_sums(.) > 0, .))

# 2. Create List
venn_list <- list(Dabbi = dabbi_taxa, Tsedey = tsedey_taxa)

# 3. Plot
ggVennDiagram(venn_list, label_alpha = 0) + 
  scale_fill_gradient(low = "white", high = "lightblue") +
  labs(title = "Shared Genera between Genotype")

```

## Venn diagram of shared species across root depths

```{r fig.width=12, fig.height=10, dpi=300}
# Install if needed: install.packages("ggVennDiagram")
library(ggVennDiagram)

# 1. Extract species lists for each Location
base_taxa <- taxa_names(subset_samples(ps_species, Depth == "Base") %>% prune_taxa(taxa_sums(.) > 0, .))
mid_section_taxa <- taxa_names(subset_samples(ps_species, Depth == "Mid-section") %>% prune_taxa(taxa_sums(.) > 0, .))
tip_taxa <- taxa_names(subset_samples(ps_species, Depth == "Tip") %>% prune_taxa(taxa_sums(.) > 0, .))

# 2. Create List
venn_list <- list(Base = base_taxa, Mid_section = mid_section_taxa, Tip = tip_taxa)

# 3. Plot
ggVennDiagram(venn_list, label_alpha = 0) + 
  scale_fill_gradient(low = "white", high = "lightblue") +
  labs(title = "Shared Genera between Genotype")

```

## Venn diagram of shared genera across genotypes

```{r fig.width=12, fig.height=10, dpi=300}
# Install if needed: install.packages("ggVennDiagram")
library(ggVennDiagram)

# 1. Extract Genus lists for each Location
dabbi_taxa <- taxa_names(subset_samples(ps_genus, Genotype == "Dabbi") %>% prune_taxa(taxa_sums(.) > 0, .))
tsedey_taxa <- taxa_names(subset_samples(ps_genus, Genotype == "Tsedey") %>% prune_taxa(taxa_sums(.) > 0, .))

# 2. Create List
venn_list <- list(Dabbi = dabbi_taxa, Tsedey = tsedey_taxa)

# 3. Plot
ggVennDiagram(venn_list, label_alpha = 0) + 
  scale_fill_gradient(low = "white", high = "lightblue") +
  labs(title = "Shared Genera between Genotype")

```

## Venn diagram of shared genera across root depths

```{r fig.width=12, fig.height=10, dpi=300}
# Install if needed: install.packages("ggVennDiagram")
library(ggVennDiagram)

# 1. Extract Genus lists for each Location
base_taxa <- taxa_names(subset_samples(ps_genus, Depth == "Base") %>% prune_taxa(taxa_sums(.) > 0, .))
mid_section_taxa <- taxa_names(subset_samples(ps_genus, Depth == "Mid-section") %>% prune_taxa(taxa_sums(.) > 0, .))
tip_taxa <- taxa_names(subset_samples(ps_genus, Depth == "Tip") %>% prune_taxa(taxa_sums(.) > 0, .))

# 2. Create List
venn_list <- list(Base = base_taxa, Mid_section = mid_section_taxa, Tip = tip_taxa)

# 3. Plot
ggVennDiagram(venn_list, label_alpha = 0) + 
  scale_fill_gradient(low = "white", high = "lightblue") +
  labs(title = "Shared Genera across Root Depths")

```

## Alpha Diversity at the genus level by genotype

```{r fig.width=12,fig.height=12, dpi=300}
# Uses ps_genus_rar (rarefied) to avoid richness inflation from unequal depth
alpha_tbl <- phyloseq::estimate_richness(
  ps_genus_rar, measures = c("Shannon", "Simpson", "Observed", "Chao1")
) %>%
  tibble::rownames_to_column("SampleID")

meta_tbl <- as(sample_data(ps_genus_rar), "data.frame") %>%
  tibble::rownames_to_column("SampleID")

df_alpha <- dplyr::left_join(meta_tbl, alpha_tbl, by = "SampleID")

df_long <- df_alpha %>%
  pivot_longer(cols = c(Shannon, Simpson, Observed, Chao1),
               names_to = "Measure", values_to = "Value")

ggplot(df_long, aes(x = Genotype, y = Value, colour = Genotype)) +
  geom_jitter(width = 0.15, height = 0, alpha = 0.8, size = 2) +
  facet_wrap(~ Measure, scales = "free_y") +
  labs(x = NULL, y = "Alpha diversity",
       title = "Alpha diversity by Genotype (rarefied)") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

# Alpha Diversity at the genus level by depth

```{r fig.width=12,fig.height=12, dpi=300}
ggplot(df_long, aes(x = Depth, y = Value, colour = Depth)) +
  geom_jitter(width = 0.15, height = 0, alpha = 0.8, size = 2) +
  facet_wrap(~ Measure, scales = "free_y") +
  labs(x = NULL, y = "Alpha diversity",
       title = "Alpha diversity by Depth (rarefied)") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

## Stats to assess differences in alpha diversity

```{r}
# ── 1. Extract metadata ───────────────────────────────────────
meta <- as(sample_data(ps_genus_rar), "data.frame") %>%
  tibble::rownames_to_column("SampleID") %>%
  dplyr::mutate(
    Genotype = factor(Genotype),
    Depth    = factor(Depth),
    Repeat   = factor(Repeat)
  )

# ── 2. Diversity metrics (rarefied) ──────────────────────────
alpha_tbl <- phyloseq::estimate_richness(
  ps_genus_rar, measures = c("Shannon", "Simpson", "Observed", "Chao1")
) %>%
  tibble::rownames_to_column("SampleID")

# ── 3. Merge ──────────────────────────────────────────────────
alpha_div <- dplyr::left_join(meta, alpha_tbl, by = "SampleID")

# ── 4. Check raw distributions ────────────────────────────────
par(mfrow = c(2, 2))
hist(alpha_div$Shannon,  main = "Shannon",  breaks = 10, col = "#66C2A5")
hist(alpha_div$Observed, main = "Observed", breaks = 10, col = "#66C2A5")
hist(alpha_div$Chao1,    main = "Chao1",    breaks = 10, col = "#66C2A5")
hist(alpha_div$Simpson,  main = "Simpson",  breaks = 10, col = "#66C2A5")
par(mfrow = c(1, 1))

# ── 5. Transformations ────────────────────────────────────────
# Five scales considered per metric:
#   raw, log (log+epsilon), sqrt, logit (Simpson only)
# NOTE: AIC is not comparable across response scales — it is used within
# each scale to choose additive vs interaction. Shapiro-Wilk and Q-Q plots
# are then used across scales to identify the best-fitting transformation.

safe_log   <- function(x) log(x + 1e-6)
safe_logit <- function(x) log((x + 1e-6) / (1 - x + 1e-6))

alpha_div <- alpha_div %>%
  dplyr::mutate(
    log_Shannon   = safe_log(Shannon),
    sqrt_Shannon  = sqrt(Shannon),
    log_Observed  = safe_log(Observed),
    sqrt_Observed = sqrt(Observed),
    log_Chao1     = safe_log(Chao1),
    sqrt_Chao1    = sqrt(Chao1),
    logit_Simpson = safe_logit(Simpson),
    sqrt_Simpson  = sqrt(Simpson)
  )

# ── 6. Fit all candidate models ───────────────────────────────
# Shannon
mod_shannon_raw_add  <- lm(Shannon      ~ Genotype + Depth, data = alpha_div)
mod_shannon_raw_int  <- lm(Shannon      ~ Genotype * Depth, data = alpha_div)
mod_shannon_log_add  <- lm(log_Shannon  ~ Genotype + Depth, data = alpha_div)
mod_shannon_log_int  <- lm(log_Shannon  ~ Genotype * Depth, data = alpha_div)
mod_shannon_sqrt_add <- lm(sqrt_Shannon ~ Genotype + Depth, data = alpha_div)
mod_shannon_sqrt_int <- lm(sqrt_Shannon ~ Genotype * Depth, data = alpha_div)

# Observed
mod_observed_raw_add  <- lm(Observed      ~ Genotype + Depth, data = alpha_div)
mod_observed_raw_int  <- lm(Observed      ~ Genotype * Depth, data = alpha_div)
mod_observed_log_add  <- lm(log_Observed  ~ Genotype + Depth, data = alpha_div)
mod_observed_log_int  <- lm(log_Observed  ~ Genotype * Depth, data = alpha_div)
mod_observed_sqrt_add <- lm(sqrt_Observed ~ Genotype + Depth, data = alpha_div)
mod_observed_sqrt_int <- lm(sqrt_Observed ~ Genotype * Depth, data = alpha_div)

# Chao1
mod_chao1_raw_add  <- lm(Chao1      ~ Genotype + Depth, data = alpha_div)
mod_chao1_raw_int  <- lm(Chao1      ~ Genotype * Depth, data = alpha_div)
mod_chao1_log_add  <- lm(log_Chao1  ~ Genotype + Depth, data = alpha_div)
mod_chao1_log_int  <- lm(log_Chao1  ~ Genotype * Depth, data = alpha_div)
mod_chao1_sqrt_add <- lm(sqrt_Chao1 ~ Genotype + Depth, data = alpha_div)
mod_chao1_sqrt_int <- lm(sqrt_Chao1 ~ Genotype * Depth, data = alpha_div)

# Simpson — logit is theoretically preferred; sqrt included for completeness
mod_simpson_raw_add   <- lm(Simpson       ~ Genotype + Depth, data = alpha_div)
mod_simpson_raw_int   <- lm(Simpson       ~ Genotype * Depth, data = alpha_div)
mod_simpson_logit_add <- lm(logit_Simpson ~ Genotype + Depth, data = alpha_div)
mod_simpson_logit_int <- lm(logit_Simpson ~ Genotype * Depth, data = alpha_div)
mod_simpson_sqrt_add  <- lm(sqrt_Simpson  ~ Genotype + Depth, data = alpha_div)
mod_simpson_sqrt_int  <- lm(sqrt_Simpson  ~ Genotype * Depth, data = alpha_div)

# ── 7. Residual diagnostics across ALL candidates ─────────────
# Run before model selection so that a model excluded by AIC but with
# better residual structure can be identified and manually preferred.

all_candidates <- list(
  shannon_raw_add  = mod_shannon_raw_add,
  shannon_raw_int  = mod_shannon_raw_int,
  shannon_log_add  = mod_shannon_log_add,
  shannon_log_int  = mod_shannon_log_int,
  shannon_sqrt_add = mod_shannon_sqrt_add,
  shannon_sqrt_int = mod_shannon_sqrt_int,

  observed_raw_add  = mod_observed_raw_add,
  observed_raw_int  = mod_observed_raw_int,
  observed_log_add  = mod_observed_log_add,
  observed_log_int  = mod_observed_log_int,
  observed_sqrt_add = mod_observed_sqrt_add,
  observed_sqrt_int = mod_observed_sqrt_int,

  chao1_raw_add  = mod_chao1_raw_add,
  chao1_raw_int  = mod_chao1_raw_int,
  chao1_log_add  = mod_chao1_log_add,
  chao1_log_int  = mod_chao1_log_int,
  chao1_sqrt_add = mod_chao1_sqrt_add,
  chao1_sqrt_int = mod_chao1_sqrt_int,

  simpson_raw_add   = mod_simpson_raw_add,
  simpson_raw_int   = mod_simpson_raw_int,
  simpson_logit_add = mod_simpson_logit_add,
  simpson_logit_int = mod_simpson_logit_int,
  simpson_sqrt_add  = mod_simpson_sqrt_add,
  simpson_sqrt_int  = mod_simpson_sqrt_int
)

# Shapiro-Wilk summary table across all candidates
shapiro_results <- data.frame(
  Model   = names(all_candidates),
  SW_W    = sapply(all_candidates, function(m) shapiro.test(residuals(m))$statistic),
  SW_p    = sapply(all_candidates, function(m) shapiro.test(residuals(m))$p.value),
  AIC_val = sapply(all_candidates, AIC)
)
shapiro_results$Normal_p05 <- shapiro_results$SW_p > 0.05

cat("── Residual normality summary (all candidates) ──\n")
print(shapiro_results, digits = 3)

# Q-Q plots for all candidates — inspect visually
for (nm in names(all_candidates)) {
  par(mfrow = c(1, 2))
  qqnorm(residuals(all_candidates[[nm]]), main = paste("Q-Q:", nm))
  qqline(residuals(all_candidates[[nm]]), col = "red")
  hist(residuals(all_candidates[[nm]]), breaks = 8,
       main = paste("Resid:", nm), col = "#66C2A5", xlab = "Residuals")
  par(mfrow = c(1, 1))
}

# ── 8. Select best model per metric ───────────────────────────
# Primary criterion: Shapiro-Wilk p > 0.05 (residuals normal)
# Secondary criterion: lowest AIC among models passing Shapiro-Wilk
# If no model passes Shapiro-Wilk, select the one with highest SW W statistic
# (closest to normal) and note this in interpretation.

select_best <- function(candidate_names, sw_table) {
  sub  <- sw_table[sw_table$Model %in% candidate_names, ]
  pass <- sub[sub$Normal_p05 == TRUE, ]
  if (nrow(pass) > 0) {
    best_name <- pass$Model[which.min(pass$AIC_val)]
    cat(sprintf("  → %s selected (passed Shapiro-Wilk, lowest AIC among passing models)\n",
                best_name))
  } else {
    best_name <- sub$Model[which.max(sub$SW_W)]
    cat(sprintf("  → %s selected (no model passed Shapiro-Wilk; best W statistic)\n",
                best_name))
  }
  all_candidates[[best_name]]
}

cat("\n── Model selection ──\n")
cat("Shannon:\n")
mod_shannon  <- select_best(grep("^shannon",  names(all_candidates), value = TRUE),
                            shapiro_results)
cat("Observed:\n")
mod_observed <- select_best(grep("^observed", names(all_candidates), value = TRUE),
                            shapiro_results)
cat("Chao1:\n")
mod_chao1    <- select_best(grep("^chao1",    names(all_candidates), value = TRUE),
                            shapiro_results)
cat("Simpson:\n")
mod_simpson  <- select_best(grep("^simpson",  names(all_candidates), value = TRUE),
                            shapiro_results)

cat("\n── Final model formulas ──\n")
cat(sprintf("Shannon:  %s\n", deparse(formula(mod_shannon))))
cat(sprintf("Observed: %s\n", deparse(formula(mod_observed))))
cat(sprintf("Chao1:    %s\n", deparse(formula(mod_chao1))))
cat(sprintf("Simpson:  %s\n", deparse(formula(mod_simpson))))

# ── 9. Full residual plots for selected models ────────────────
par(mfrow = c(2, 2))
plot(mod_shannon);  par(mfrow = c(2, 2))
plot(mod_observed); par(mfrow = c(2, 2))
plot(mod_chao1);    par(mfrow = c(2, 2))
plot(mod_simpson)
par(mfrow = c(1, 1))

# ── 10. ANOVA tables ──────────────────────────────────────────
cat("\n── Shannon ANOVA ──\n");  print(Anova(mod_shannon,  type = "II"))
cat("\n── Observed ANOVA ──\n"); print(Anova(mod_observed, type = "II"))
cat("\n── Chao1 ANOVA ──\n");    print(Anova(mod_chao1,    type = "II"))
cat("\n── Simpson ANOVA ──\n");  print(Anova(mod_simpson,  type = "II"))

# ── 11. Effect sizes ──────────────────────────────────────────
eta_squared(Anova(mod_shannon,  type = "II"), partial = FALSE, ci = 0.90)
eta_squared(Anova(mod_observed, type = "II"), partial = FALSE, ci = 0.90)
eta_squared(Anova(mod_chao1,    type = "II"), partial = FALSE, ci = 0.90)
eta_squared(Anova(mod_simpson,  type = "II"), partial = FALSE, ci = 0.90)

# ── 12. Pairwise comparisons ──────────────────────────────────
pairs(emmeans(mod_shannon,  ~ Depth),    adjust = "sidak")
pairs(emmeans(mod_shannon,  ~ Genotype), adjust = "sidak")
pairs(emmeans(mod_observed, ~ Depth),    adjust = "sidak")
pairs(emmeans(mod_observed, ~ Genotype), adjust = "sidak")
pairs(emmeans(mod_chao1,    ~ Depth),    adjust = "sidak")
pairs(emmeans(mod_chao1,    ~ Genotype), adjust = "sidak")
pairs(emmeans(mod_simpson,  ~ Depth),    adjust = "sidak")
pairs(emmeans(mod_simpson,  ~ Genotype), adjust = "sidak")

# ── 13. Interaction cell contrasts ────────────────────────────
for (metric in c("Shannon", "Observed", "Chao1", "Simpson")) {
  mod <- get(paste0("mod_", tolower(metric)))
  if (grepl("\\*", deparse(formula(mod)))) {
    cat(sprintf("\n── %s: Depth within each Genotype ──\n", metric))
    print(pairs(emmeans(mod, ~ Depth    | Genotype), adjust = "sidak"))
    cat(sprintf("\n── %s: Genotype at each Depth ──\n", metric))
    print(pairs(emmeans(mod, ~ Genotype | Depth),    adjust = "sidak"))
  }
}

# Normality of residuals was assessed visually via Q-Q plots and confirmed with Shapiro-Wilk tests; where 
# Shapiro-Wilk indicated significant departures (Chao1, Simpson), residual Q-Q plots showed only minor tail 
# deviations consistent with ANOVA robustness at the study's sample size (n = 22).
```

## Differential abundance - which taxa drive differences

```{r}
# ── Setup ─────────────────────────────────────────────────────
otu_mat_genus <- as(phyloseq::otu_table(ps_genus), "matrix")
if (!phyloseq::taxa_are_rows(ps_genus)) otu_mat_genus <- t(otu_mat_genus)
otu_mat_genus <- t(otu_mat_genus)  # vegan needs samples as rows

meta_simper <- as(phyloseq::sample_data(ps_genus), "data.frame")
meta_simper <- meta_simper[rownames(otu_mat_genus), ]

# ── SIMPER by depth ───────────────────────────────────────────
set.seed(123)
simper_depth <- vegan::simper(otu_mat_genus, 
                               group = meta_simper$Depth,
                               permutations = 999)
summary(simper_depth)

# ── SIMPER by genotype ────────────────────────────────────────
set.seed(123)
simper_geno <- vegan::simper(otu_mat_genus, 
                              group = meta_simper$Genotype,
                              permutations = 999)
summary(simper_geno)

```

## PERMANOVA analysis on whole phyloseq data at genus level

```{r}
# ── Beta Diversity & PERMANOVA ────────────────────────────────────────────
# All models use by = "margin" (marginal/Type III-equivalent) so that each
# term is tested after accounting for all other terms in the model.
# This is the correct approach for a balanced two-factor design and ensures
# results are directly comparable across additive and interaction models.
# Note: adonis2 with by = "margin" tests the interaction term by comparing
# the full model against the model without that term — this is valid but
# consumes degrees of freedom, so interpret the interaction with caution
# given n = 22.

# ── 1. Compute distances ──────────────────────────────────────────────────
braydist      <- phyloseq::distance(ps_genus,     method = "bray")
aitchisondist <- phyloseq::distance(ps_genus_clr, method = "euclidean")

meta <- as(sample_data(ps_genus), "data.frame") %>%
  dplyr::mutate(Genotype  = factor(Genotype),
                Depth     = factor(Depth),
                GenoDepth = interaction(Genotype, Depth))
meta <- meta[labels(braydist), ]

# ── 2. Betadisper ─────────────────────────────────────────────────────────
cat("Betadisper — Genotype\n")
print(permutest(betadisper(braydist, meta$Genotype), permutations = 999))

cat("\nBetadisper — Depth\n")
print(permutest(betadisper(braydist, meta$Depth), permutations = 999))

cat("\nBetadisper — Genotype:Depth\n")
print(permutest(betadisper(braydist, meta$GenoDepth), permutations = 999))

# ── 3. PERMANOVA additive (primary) ───────────────────────────────────────
cat("\nPERMANOVA Bray-Curtis — Additive\n")
set.seed(123)
print(adonis2(braydist ~ Genotype + Depth, data = meta,
              permutations = 999, by = "margin"))

cat("\nPERMANOVA Aitchison — Additive\n")
set.seed(123)
print(adonis2(aitchisondist ~ Genotype + Depth, data = meta,
              permutations = 999, by = "margin"))

# ── 4. PERMANOVA interaction (exploratory) ────────────────────────────────
cat("\nPERMANOVA Bray-Curtis — Interaction (exploratory)\n")
set.seed(123)
print(adonis2(braydist ~ Genotype * Depth, data = meta,
              permutations = 999, by = "margin"))

cat("\nPERMANOVA Aitchison — Interaction (exploratory)\n")
set.seed(123)
print(adonis2(aitchisondist ~ Genotype * Depth, data = meta,
              permutations = 999, by = "margin"))

# ── 5. Pairwise depth comparisons with Bonferroni correction ─────────────
bray_mat <- as.matrix(braydist)

run_pairwise <- function(d1, d2) {
  m <- meta[meta$Depth %in% c(d1, d2), ]
  m$Depth <- factor(m$Depth)
  d <- as.dist(bray_mat[rownames(m), rownames(m)])
  set.seed(123)
  res <- adonis2(d ~ Depth, data = m, permutations = 999, by = "margin")
  res["Depth", "Pr(>F)"]
}

p_raw <- c(
  "Base vs Mid-section" = run_pairwise("Base", "Mid-section"),
  "Base vs Tip"         = run_pairwise("Base", "Tip"),
  "Mid-section vs Tip"  = run_pairwise("Mid-section", "Tip")
)
p_bonf <- pmin(p_raw * length(p_raw), 1)

cat("\nPairwise depth PERMANOVA (Bonferroni corrected):\n")
print(data.frame(
  p_raw        = round(p_raw,  4),
  p_bonferroni = round(p_bonf, 4),
  significant  = p_bonf < 0.05
))

# ── 6. Genotype PERMANOVA within each depth ───────────────────────────────
for (depth_level in levels(meta$Depth)) {
  cat(sprintf("\nGenotype PERMANOVA at depth: %s\n", depth_level))
  ps_sub   <- phyloseq::subset_samples(ps_genus, Depth == depth_level)
  bray_sub <- phyloseq::distance(ps_sub, method = "bray")
  meta_sub <- as(sample_data(ps_sub), "data.frame") %>%
    dplyr::mutate(Genotype = factor(Genotype))
  meta_sub <- meta_sub[labels(bray_sub), ]
  print(table(meta_sub$Genotype))
  if (any(table(meta_sub$Genotype) < 2)) { cat("Skipping.\n"); next }
  set.seed(123)
  print(adonis2(bray_sub ~ Genotype, data = meta_sub,
                permutations = 999, by = "margin"))
}

```

## Identifying taxa characteristic of a specific genotype or depth

```{r}
library(indicspecies)

otu_df <- as.data.frame(phyloseq::otu_table(ps_species))
if (!phyloseq::taxa_are_rows(ps_species)) otu_df <- t(otu_df)

meta_ind <- as(phyloseq::sample_data(ps_species), "data.frame")
meta_ind <- meta_ind[colnames(otu_df), ]

# ── By genotype ───────────────────────────────────────────────
indval_genotype <- indicspecies::multipatt(
  t(otu_df),
  meta_ind$Genotype,
  control = permute::how(nperm = 999)
)
summary(indval_genotype, alpha = 0.05)

# ── Genotype indicator species taxonomy ───────────────────────
indicator_taxa_genotype <- phyloseq::tax_table(ps_species) %>%
  as.data.frame() %>%
  tibble::rownames_to_column("OTU") %>%
  dplyr::filter(OTU %in% c("OTU179", "OTU66"))

cat("── Genotype indicator species taxonomy ──\n")
print(indicator_taxa_genotype)

# ── By depth ──────────────────────────────────────────────────
indval_depth <- indicspecies::multipatt(
  t(otu_df),
  meta_ind$Depth,
  control = permute::how(nperm = 999)
)
summary(indval_depth, alpha = 0.05)

# ── Depth indicator species taxonomy ─────────────────────────
indicator_taxa_depth <- phyloseq::tax_table(ps_species) %>%
  as.data.frame() %>%
  tibble::rownames_to_column("OTU") %>%
  dplyr::filter(OTU %in% c("OTU227", "OTU42"))

cat("── Depth indicator species taxonomy ──\n")
print(indicator_taxa_depth)

# NOTE: Indicator values computed on raw (non-rarefied) data — taxa abundant
# in high-read samples may have inflated IndVal scores. Interpret accordingly.
# Note that the indiciator species identified were assessed on the raw (non-rarefied) data which risks inflating the indicator values for taxa that are abundant in high-read samples - take this into account.
```

## Indicator spcies genotype_depth

```{r}
library(indicspecies)

# ─────────────────────────────────────────────────────────────
# STEP 1: Prepare species-level data
# ─────────────────────────────────────────────────────────────
otu_df <- as.data.frame(phyloseq::otu_table(ps_species))
if (!phyloseq::taxa_are_rows(ps_species)) {
  otu_df <- t(otu_df)  # Rows = samples, columns = OTUs
}

meta_ind <- as(phyloseq::sample_data(ps_species), "data.frame")
meta_ind <- meta_ind[colnames(otu_df), ]  # Match sample order
meta_ind$genotype_depth <- paste(meta_ind$Genotype, meta_ind$Depth, sep = "_")

# ─────────────────────────────────────────────────────────────
# STEP 2: Indicator species analysis
# ─────────────────────────────────────────────────────────────
results_list <- list()

for (group_name in unique(meta_ind$genotype_depth)) {
  cat(sprintf("\n── %s ──\n", group_name))
  
  # Binary grouping: THIS group vs all others
  group_binary <- meta_ind$genotype_depth == group_name
  
  # IndVal analysis
  indval_single <- indicspecies::multipatt(
    t(otu_df),
    group_binary,
    control = permute::how(nperm = 999)
  )
  
  # All significant OTUs (p < 0.05)
  sig_indicators <- summary(indval_single, alpha = 0.05)$sign
  
  if (is.null(sig_indicators) || nrow(sig_indicators) == 0) {
    cat("  → No significant OTUs\n")
    next
  }
  
  # Store ALL significant results for this group
  results_list[[group_name]] <- sig_indicators
  cat(sprintf("  → %d significant OTUs (p < 0.05)\n", nrow(sig_indicators)))
  print(head(sig_indicators[, c("stat", "p.value")], 3))
}

# ─────────────────────────────────────────────────────────────
# STEP 3: Summary table of OTUs per group
# ─────────────────────────────────────────────────────────────
cat("\n── SUMMARY: OTUs per genotype-depth ──\n")
otu_summary <- data.frame(
  Group = names(results_list),
  N_OTUs = sapply(results_list, nrow),
  Top_OTUs = sapply(results_list, function(x) {
    paste(head(rownames(x[order(x$stat, decreasing = TRUE), ]), 2), collapse = ", ")
  })
)
print(otu_summary)

# ── OUTPUT: Save OTU list for taxonomy lookup ───────────────────────────
all_sig_otus <- unique(unlist(lapply(results_list, rownames)))
cat("\n── ALL SIGNIFICANT OTUs (copy to taxonomy chunk) ──\n")
cat(paste('"', all_sig_otus, '"', collapse = ", "), "\n")
cat(sprintf("\nTotal unique OTUs: %d\n", length(all_sig_otus)))

cat("\nNOTE: Run taxonomy lookup on these OTUs in separate chunk.\n")
```

## Finding identity of OTUs identified

```{r}
# ── FIXED: Updated with your ACTUAL significant OTUs ─────────────────────
# These 17 OTUs were flagged by multipatt() (p < 0.05)

significant_otus <- c("OTU126", "OTU179", "OTU309", "OTU187",  # Dabbi_Mid-section
                      "OTU23", "OTU142", "OTU88", "OTU252", "OTU183", "OTU227",  # Dabbi_Tip
                      "OTU66",                               # Tsedey_Base  
                      "OTU327", "OTU219", "OTU324", "OTU42") # Tsedey_Tip

# ── Safe taxonomy lookup (no more subscript error) ───────────────────────
cat("── Checking taxonomy match ──\n")
tax_table_full <- phyloseq::tax_table(ps_species)
available_otus <- rownames(tax_table_full)

matching_otus <- intersect(significant_otus, available_otus)
cat(sprintf("Found %d/%d OTUs in taxonomy table\n", 
            length(matching_otus), length(significant_otus)))

if (length(matching_otus) > 0) {
  # Extract taxonomy for matching OTUs only
  tax_table_subset <- tax_table_full[matching_otus, ]
  
  otu_taxonomy <- as.data.frame(tax_table_subset) %>%
    tibble::rownames_to_column("OTU") %>%
    dplyr::mutate(
      Species_name = case_when(
        !is.na(Species) & Species != "" ~ Species,
        !is.na(Genus) & Genus != "" ~ paste(Genus, "sp."),
        TRUE ~ "Unclassified"
      )
    ) %>%
    dplyr::arrange(OTU)
  
  cat("\n── IDENTITIES OF SIGNIFICANT INDICATOR OTUs ──\n")
  print(otu_taxonomy)
} else {
  cat("No matching OTUs found. Check:\n")
  cat("Available OTUs:", paste(head(available_otus, 5), collapse = ", "), "...\n")
  cat("Your OTUs:   ", paste(head(significant_otus, 5), collapse = ", "), "...\n")
}

# ── Biological interpretation (group associations) ───────────────────────
cat("\n── BIOLOGICAL INTERPRETATION ──\n")

# Dabbi_Mid-section (4 specialists)
dabbi_mid <- c("OTU126", "OTU179", "OTU309", "OTU187")
cat("**Dabbi_Mid-section specialists:**\n")
if (exists("otu_taxonomy")) {
  print(otu_taxonomy[otu_taxonomy$OTU %in% dabbi_mid, "Species_name"])
}

# Dabbi_Tip (7 specialists)  
dabbi_tip <- c("OTU23", "OTU142", "OTU88", "OTU252", "OTU179", "OTU183", "OTU227")
cat("\n**Dabbi_Tip specialists:**\n")
if (exists("otu_taxonomy")) {
  print(otu_taxonomy[otu_taxonomy$OTU %in% dabbi_tip, "Species_name"])
}

# Tsedey_Base
tsedey_base <- "OTU66"
cat("\n**Tsedey_Base:**\n")
if (exists("otu_taxonomy")) {
  print(otu_taxonomy[otu_taxonomy$OTU %in% tsedey_base, "Species_name"])
}

# Tsedey_Tip
tsedey_tip <- c("OTU327", "OTU219", "OTU324", "OTU42")
cat("\n**Tsedey_Tip:**\n")
if (exists("otu_taxonomy")) {
  print(otu_taxonomy[otu_taxonomy$OTU %in% tsedey_tip, "Species_name"])
}
```


## Rarefaction curve to see if sequencing depth was sufficient to find all species (by depth)

```{r fig.width=10, fig.height=8, dpi=300}
# ── 1. Extract OTU table ──────────────────────────────────────
otu_mat_counts <- as(phyloseq::otu_table(ps), "matrix")
if (phyloseq::taxa_are_rows(ps)) otu_mat_counts <- t(otu_mat_counts)

# ── 2. Build palette dynamically to match number of depth levels ─
depth_levels  <- levels(factor(phyloseq::sample_data(ps)$Depth))
n_depths      <- length(depth_levels)
depth_palette <- RColorBrewer::brewer.pal(max(3, n_depths), "Set2")[1:n_depths]

# ── 3. Map colours to samples ────────────────────────────────
sample_colors <- depth_palette[as.numeric(factor(phyloseq::sample_data(ps)$Depth))]

# ── 4. Plot rarefaction curves ────────────────────────────────
vegan::rarecurve(
  otu_mat_counts,
  step  = 20,
  col   = sample_colors,
  lwd   = 1.5,
  label = FALSE,
  main  = "Rarefaction Curves by Depth",
  xlab  = "Sequencing Depth (Reads)",
  ylab  = "Observed Richness"
)

# ── 5. Legend ─────────────────────────────────────────────────
legend(
  "bottomright",
  legend  = depth_levels,
  col     = depth_palette,
  lty     = 1,
  lwd     = 3,
  cex     = 0.8,
  title   = "Depth",
  box.lwd = 0
)

```

# rarefaction by gentoype

```{r fig.width=10, fig.height=8, dpi=300}
# ── 1. Extract OTU table ──────────────────────────────────────
otu_mat_counts <- as(phyloseq::otu_table(ps), "matrix")
if (phyloseq::taxa_are_rows(ps)) otu_mat_counts <- t(otu_mat_counts)

# ── 2. Build palette dynamically ─────────────────────────────
geno_levels   <- levels(factor(phyloseq::sample_data(ps)$Genotype))
n_geno        <- length(geno_levels)
geno_palette  <- RColorBrewer::brewer.pal(max(3, n_geno), "Set2")[1:n_geno]

# ── 3. Map colours to samples ─────────────────────────────────
sample_colors <- geno_palette[as.numeric(factor(phyloseq::sample_data(ps)$Genotype))]

# ── 4. Plot ───────────────────────────────────────────────────
vegan::rarecurve(
  otu_mat_counts,
  step  = 20,
  col   = sample_colors,
  lwd   = 1.5,
  label = FALSE,
  main  = "Rarefaction Curves by Genotype",
  xlab  = "Sequencing Depth (Reads)",
  ylab  = "Observed Richness"
)

# ── 5. Legend ─────────────────────────────────────────────────
legend(
  "bottomright",
  legend  = geno_levels,
  col     = geno_palette,
  lty     = 1,
  lwd     = 3,
  cex     = 0.8,
  title   = "Genotype",
  box.lwd = 0
)

```
