---
title: "RSA_day16_agar"
author: "Alessandra Learmount"
format:
  pdf:
    fig-pos: "H"
toc: true
date: "2026-03-03"
---
# Root Architecture Analysis of Day 16 Roots Grown on Agar

## Summary
This MD was used to assess 10 different root system architecture traits on 16-17 day old roots grown on agar in square plates, with 1-4 plants per plate. Roots were scanned directly in the square plates using an Epson V850 flatbed scanner.


Loading librairies, importing dataset and obtaining day, batch and genotype

```{r}
library(tidyverse)
library(moments)
library(lme4)
library(emmeans)
library(multcomp)
library(multcompView)
library(lmerTest)
library(ggplot2)
library(dplyr)
library(stringr)
library(car)
library(ggpubr)
library(patchwork)
library(performance)
library(ggResidpanel)
library(glmmTMB)
library(DHARMa)
library(performance)
library(cld)

df <- read.csv("~/Desktop/RStudio/teff_rotation_apr26/features_day1617.csv")

# extract relevant data from teh csv file
df <- df %>%
  mutate(
    ID            = str_remove(File.Name, "_seg\\.png"),
    Day           = str_extract(File.Name, "D\\d+"),
    Batch         = str_extract(File.Name, "B\\d+"),
    Genotype      = str_extract(File.Name, "(?<=_B\\d_)[A-Z]+(?=P)"),
    Plate         = str_extract(File.Name, "(?<=P)\\d+(?=S)"),
    Seed          = str_extract(File.Name, "(?<=S)\\d+(?=_seg)"),
    Genotype_Name = case_when(
      Genotype == "A" ~ "Addisie",
      Genotype == "B" ~ "Beten",
      Genotype == "D" ~ "Dabbi",
      Genotype == "K" ~ "Karadebi",
      Genotype == "M" ~ "Manyi",
      Genotype == "T" ~ "Tsedey",
      TRUE ~ NA_character_
    )
  ) %>%
   # Explicitly coerce grouping variables to factors
  mutate(
    Genotype = factor(Genotype_Name),
    Batch    = factor(Batch),
    Plate    = factor(Plate)
  ) %>%
  # Drop any rows where key variables failed to parse
  filter(!is.na(Genotype), !is.na(Batch), !is.na(Plate))


head(df)
```

Overall visualisation and defining traits - bix panel boxplot

```{r, fig.width=20, fig.height=12, dpi=300}
key_traits <- c("Total.Root.Length.mm", "Median.Number.of.Roots", "Maximum.Number.of.Roots", "Maximum.Width.mm", "Depth.mm", "Number.of.Root.Tips",
                "Average.Diameter.mm", "Maximum.Diameter.mm", "Width.to.Depth.Ratio",
                "Network.Area.mm2", "Convex.Area.mm2", "Perimeter.mm", "Volume.mm3", "Surface.Area.mm2", "Average.Root.Orientation.deg",
                "Steep.Angle.Frequency", "Shallow.Angle.Frequency", "Medium.Angle.Frequency", "Solidity", "Lower.Root.Area.mm2")

df_long <- df %>% pivot_longer(all_of(key_traits), names_to = "Trait", values_to = "Value")

head(df_long)

ggplot(df_long, aes(x = Genotype, y = Value, fill = Genotype)) +
  geom_boxplot(alpha = 0.6) +
  geom_jitter(aes(colour = Batch), width = 0.15, size = 1.5) +
  scale_fill_brewer(palette = "Set2") +
  scale_colour_brewer(palette = "Set2") +
  facet_wrap(~Trait, scales = "free_y") +
  theme_bw()

```

Overall, box plot suggests broadly simlar phenotypes across the species so will pick out individual traits to see if any significant differences do arise. 

### Number of Root Tips

Assessing each factor individually

-   Assess type of data

-   Decide best model

-   Check assumptions

-   Stats tests

-   Plots

```{r, fig.width=16, fig.height=12, dpi=400}
# ============================================================
#  Number of Root Tips Analysis
# ============================================================

# ── 1. Inspect distribution ───────────────────────────────────
# Root tips is a non-negative integer count — must use a count
# distribution (Poisson or NB), not Gaussian.
summary(df$Number.of.Root.Tips)
cat("Var/Mean ratio:", round(var(df$Number.of.Root.Tips) /
      mean(df$Number.of.Root.Tips), 3),
    "(>1 suggests overdispersion)\n")

# Histogram to visualise shape and identify skew or zero-inflation
ggplot(df, aes(x = Number.of.Root.Tips)) +
  geom_histogram(binwidth = 3, fill = "#66C2A5", colour = "white") +
  geom_vline(xintercept = mean(df$Number.of.Root.Tips),
             linetype = "dashed", colour = "red") +
  labs(x = "Number of root tips", y = "Count") +
  theme_classic(base_size = 14)

# ── 2. Fit Poisson to formally confirm overdispersion ─────────
# We fit Poisson first to formally test whether its assumption
# (variance = mean) holds. If DHARMa flags overdispersion
# (p < 0.05, ratio >> 1), we reject Poisson and move to NB.
# Random effects: seeds nested in plates nested in batches —
# ignoring this clustering would pseudoreplicate as plants on
# the same plate share physical and nutritional environment.
tips_tmb_pois <- glmmTMB(
  Number.of.Root.Tips ~ Genotype + (1 | Batch/Plate/Seed),
  data = df, family = poisson
)

sim_pois <- simulateResiduals(tips_tmb_pois, n = 1000)
plot(sim_pois)
testDispersion(sim_pois)   # p < 0.05 → overdispersed → try NB to see if there is a better fit

# the residual plots show from outlier test the deviation is significant so we will try NB

# ── 3. Compare nb2 vs nb1 ─────────────────────────────────────
# Two NB parameterisations:
#   nb2: variance = mu + mu²/theta — quadratic, most common for
#        biological count data where variance scales with mean²
#   nb1: variance = phi*mu — linear, less common
# Select using AIC (lower = better) and DHARMa visual diagnostics.

tips_tmb_nb2 <- glmmTMB(
  Number.of.Root.Tips ~ Genotype + (1 | Batch/Plate/Seed),
  data = df, family = nbinom2
)

tips_tmb_nb1 <- glmmTMB(
  Number.of.Root.Tips ~ Genotype + (1 | Batch/Plate/Seed),
  data = df, family = nbinom1
)

# AIC comparison — lower = better fit penalised for complexity
AIC(tips_tmb_pois, tips_tmb_nb1, tips_tmb_nb2)

# find that NB2 is better than poisson and NB1 fails so will continue with NB2

# DHARMa residuals for each NB model — choose the one with
# better QQ alignment and flatter residual vs fitted panel
sim_nb2 <- simulateResiduals(tips_tmb_nb2, n = 1000)
plot(sim_nb2)

# ── 4. Full assumption checks on chosen model (nb2) ───────────
# testUniformity: residuals should be uniformly distributed (p > 0.05)
# testDispersion: ratio ≈ 1.0 confirms variance correctly specified
# testZeroInflation: checks if zeros are over-represented beyond NB
# testOutliers: flags any observations with extreme residuals
testUniformity(sim_nb2)
testDispersion(sim_nb2)
testZeroInflation(sim_nb2)
testOutliers(sim_nb2)

# residual plots show no significant deviation

# ── 5. Random effects structure ───────────────────────────────
VarCorr(tips_tmb_nb2)
check_singularity(tips_tmb_nb2) 

# ── 7. Test Genotype effect + pairwise comparisons ────────────
# emmeans gives estimated marginal means on the count scale
# (type = "response" back-transforms from log link).
# sidak HSD controls family-wise error rate across all pairs.
car::Anova(tips_tmb_nb2, type = "III")

emm <- emmeans(tips_tmb_nb2, ~ Genotype, type = "response")
pairs(emm, adjust = "sidak")

cld_df <- as.data.frame(cld(emm, adjust = "sidak",
                             Letters = letters, alpha = 0.05)) %>%
  mutate(.group = trimws(.group))

names(cld_df)[names(cld_df) %in% c("asymp.LCL", "lower.CL")] <- "LCL"
names(cld_df)[names(cld_df) %in% c("asymp.UCL", "upper.CL")] <- "UCL"

n_df <- df %>%
  count(Genotype) %>%
  mutate(label = paste0("n=", n))


# Effect size

# Cohen's d = difference in means / pooled SD
# Interpretation: 0.2 = small, 0.5 = medium, 0.8 = large

emm_tips <- emmeans(tips_tmb_nb2, ~ Genotype)

eff_tips <- eff_size(
  emm_tips,
  sigma = sigma(tips_tmb_nb2),   # residual SD from model
  edf   = df.residual(tips_tmb_nb2)
)
print(eff_tips)

# all effect sizes are small 

# Summary stats needed for bar chart
bar_df <- df %>%
  group_by(Genotype) %>%
  summarise(mean_tips = mean(Number.of.Root.Tips),
            se        = sd(Number.of.Root.Tips) / sqrt(n()),
            .groups   = "drop")


# ── 8. Plot 1 — Boxplot + jitter ─────────────────────────────
# Boxplot shows full data distribution; jitter overlays raw points.
# CLD letters indicate sidak groupings; shared letters = ns.
# n labels below x-axis show sample sizes per genotype.
y_ceil <- max(df$Number.of.Root.Tips) * 1.2
y_label <- y_ceil * 0.86

p1 <- ggplot(df, aes(x = Genotype, y = Number.of.Root.Tips, fill = Genotype)) +

  geom_boxplot(
    alpha         = 0.7,
    outlier.shape = NA,
    width         = 0.5,
    colour        = "black",
    linewidth     = 0.5
  ) +

  geom_jitter(
    width  = 0.15,
    size   = 1.8,
    alpha  = 0.55,
    colour = "black"
  ) +

  geom_text(
    data        = cld_df,
    aes(x = Genotype, label = .group),
    y           = y_label,
    inherit.aes = FALSE,
    size        = 9,
    fontface    = "bold",
    colour      = "grey20"
  ) +

  geom_text(
    data        = n_df,
    aes(x = Genotype, label = label),
    y           = -Inf,
    vjust       = 4.2,
    inherit.aes = FALSE,
    size        = 6,
    colour      = "grey40"
  ) +

  scale_fill_brewer(palette = "Set2") +

  labs(
    x       = "Genotype",
    y       = "Number of root tips",
    caption = "Shared letters indicate no significant difference (sidak HSD, p > 0.05; glmmTMB, nbinom2). Bars show raw means ± SE."
  ) +

  theme_classic(base_size = 14) +
  theme(
    legend.position = "none",
    panel.grid      = element_blank(),
    axis.line       = element_line(colour = "black", linewidth = 0.6),
    axis.ticks      = element_line(colour = "black", linewidth = 0.5),
    axis.title      = element_text(size = 26, face = "bold"),
    axis.title.x    = element_text(size = 26, face = "bold", margin = margin(t = 24)),
    axis.text       = element_text(size = 22, colour = "black"),
    axis.text.x     = element_text(face = "italic", margin = margin(t = 15)),
    plot.caption    = element_text(size = 12, colour = "grey40", hjust = 0, margin = margin(t = 10)),
    plot.margin     = margin(t = 25, r = 25, b = 25, l = 25)
  )

p1 + coord_cartesian(clip = "off")

# ── 9. Plot 2 — Bar chart (mean ± SE) ────────────────────────
# Bar chart shows mean ± 1 SE per genotype with raw data overlaid.
# SE bars give a sense of within-genotype variability.
# Raw points overlaid for transparency — avoids hiding the spread.
# CLD letters from the mixed model placed above bars.

p2 <- ggplot(bar_df, aes(x = Genotype, y = mean_tips, fill = Genotype)) +

  geom_col(
    colour    = "black",
    linewidth = 0.3,
    width     = 0.5,
    alpha     = 0.8555
  ) +

  geom_errorbar(
    aes(ymin = mean_tips - se, ymax = mean_tips + se),
    width     = 0.2,
    linewidth = 0.3,
    colour    = "black"
  ) +

  # Raw data overlaid for transparency
  geom_jitter(
    data   = df,
    aes(x = Genotype, y = Number.of.Root.Tips),
    width  = 0.15,
    size   = 1.3,
    alpha  = 0.55,
    colour = "black",
    inherit.aes = FALSE
  ) +

  geom_text(
    data        = cld_df,
    aes(x = Genotype, label = .group),
    y           = max(df$Number.of.Root.Tips, na.rm = TRUE) * 1.02,
    inherit.aes = FALSE,
    size        = 5,
    fontface    = "bold",
    colour      = "grey20"
  ) +

  geom_text(
    data        = n_df,
    aes(x = Genotype, label = label),
    y           = -Inf,
    vjust       = 4.2,
    inherit.aes = FALSE,
    size        = 3.5,
    colour      = "grey40"
  ) +

  scale_fill_brewer(palette = "Set2") +

  labs(
    x       = "Genotype",
    y       = "Number of root tips",
    caption = "Bars = mean ± 1 SE; points = raw data\nShared letters indicate no significant difference (sidak HSD, p > 0.05; glmmTMB, nbinom2)"
  ) +

  theme_classic(base_size = 18) +
  theme(
    legend.position = "none",
    panel.grid      = element_blank(),
    axis.line       = element_line(colour = "black", linewidth = 0.6),
    axis.ticks      = element_line(colour = "black", linewidth = 0.5),
    axis.title      = element_text(size = 16, face = "bold"),
    axis.text       = element_text(size = 16, colour = "black"),
    axis.text.x     = element_text(face = "italic", margin = margin(t = 3)),
    plot.caption    = element_text(size = 12, colour = "grey40", hjust = 0),
    plot.margin     = margin(t = 10, r = 15, b = 30, l = 10)
  )

p2 + coord_cartesian(clip = "off")

```

### Average Root Diameter

```{r, fig.width=16, fig.height=12, dpi=300}

df %>%
  group_by(Genotype) %>%
  summarise(Mean     = round(mean(Average.Diameter.mm), 2),
            Variance = round(var(Average.Diameter.mm),  2),
            Var_Mean = round(var(Average.Diameter.mm) /
                               mean(Average.Diameter.mm), 2))


# Histogram
ggplot(df, aes(x = Average.Diameter.mm)) +
  geom_histogram(bins = 15, fill = "#66C2A5", colour = "white") +
  labs(title = "Distribution of Average Root Diameter",
       x = "Average Diameter (mm)", y = "Count") +
  theme_bw()

# ── Original model (Batch/Plate/Seed nested shorthand) ───────
# R expands this into THREE separate variance components:
# (1|Batch) + (1|Batch:Plate) + (1|Batch:Plate:Seed)

diam_tmb_full <- glmmTMB(
  Average.Diameter.mm ~ Genotype + (1 | Batch/Plate/Seed),
  data = df, family = gaussian()
)

# Inspect variance components
VarCorr(diam_tmb_full)
check_singularity(diam_tmb_full) 


# Because Plate and Batch variances collapse to ~0, the Hessian
# matrix becomes singular and all SEs are uncomputable:

# ── Fit reduced models for best fit ─────────────────────────────────────────────────
diam_tmb <- glmmTMB(
  Average.Diameter.mm ~ Genotype + (1 | Batch/Plate),
  data = df, family = gaussian()
)

VarCorr(diam_tmb)
check_singularity(diam_tmb) 

# ── Assumption checks ─────────────────────────────────────────
sim_res <- simulateResiduals(diam_tmb, n = 1000)
plot(sim_res)                                  # QQ + residuals vs fitted
testUniformity(sim_res)                        # p > 0.05 = Gaussian appropriate
testDispersion(sim_res)                        # ratio ≈ 1.0 = variance correct
plotResiduals(sim_res, form = df$Genotype)     # no patterns per genotype

# residual fits are all good so happy to continue with this model 

# ── Test overall Genotype effect ──────────────────────────────
car::Anova(diam_tmb, type = "III")

# ── Pairwise comparisons + CLD ────────────────────────────────
emm       <- emmeans(diam_tmb, ~ Genotype)
pairs(emm, adjust = "sidak")

cld_df <- as.data.frame(cld(emm, adjust = "sidak",
                             Letters = letters, alpha = 0.05)) %>%
  mutate(.group = trimws(.group))


# Cohen's d = difference in means / pooled SD
# Interpretation: 0.2 = small, 0.5 = medium, 0.8 = large

emm_diam <- emmeans(diam_tmb, ~ Genotype)

eff_diam <- eff_size(
  emm_diam,
  sigma = sigma(diam_tmb),   # residual SD from model
  edf   = df.residual(diam_tmb)
)
print(eff_diam)

# most effect sizes are relatively small

# ── Sample size labels ────────────────────────────────────────
n_df <- df %>%
  count(Genotype) %>%
  mutate(label = paste0("n=", n))

# ── CLD letter position (above tallest box) ───────────────────

y_ceil <- max(df$Average.Diameter.mm) * 1.2
y_label <- y_ceil * 0.85

# ── Plot 1: Boxplot + jitter (mirrors your template exactly) ──
p1 <- ggplot(df, aes(x = Genotype, y = Average.Diameter.mm, fill = Genotype)) +

  geom_boxplot(
    alpha         = 0.7,
    outlier.shape = NA,
    width         = 0.5,
    colour        = "black",
    linewidth     = 0.5
  ) +

  geom_jitter(
    width  = 0.15,
    size   = 1.8,
    alpha  = 0.55,
    colour = "black"
  ) +

  geom_text(
    data        = cld_df,
    aes(x = Genotype, label = .group),
    y           = y_label,
    inherit.aes = FALSE,
    size        = 8,
    fontface    = "bold",
    colour      = "grey20"
  ) +

  geom_text(
    data        = n_df,
    aes(x = Genotype, label = label),
    y           = -Inf,
    vjust       = 4.2,
    inherit.aes = FALSE,
    size        = 6,
    colour      = "grey40"
  ) +

  scale_fill_brewer(palette = "Set2") +

  labs(
    x       = "Genotype",
    y       = "Average root diameter (mm)",
    caption = "Shared letters indicate no significant difference (sidak HSD, p > 0.05; glmmTMB, Gaussian)"
  ) +

  theme_classic(base_size = 14) +
  theme(
    legend.position = "none",
    panel.grid      = element_blank(),
    axis.line       = element_line(colour = "black", linewidth = 0.6),
    axis.ticks      = element_line(colour = "black", linewidth = 0.5),
    axis.title      = element_text(size = 26, face = "bold"),
    axis.title.x    = element_text(size = 26, face = "bold", margin = margin(t = 24)),
    axis.text       = element_text(size = 22, colour = "black"),
    axis.text.x     = element_text(face = "italic", margin = margin(t = 15)),
    plot.caption    = element_text(size = 12, colour = "grey40", hjust = 0, margin = margin(t = 10)),
    plot.margin     = margin(t = 25, r = 25, b = 25, l = 25)
  )


p1 + coord_cartesian(clip = "off")

```

### Convex Area

```{r, fig.width=16, fig.height=12, dpi=300}

# ── 1. Inspect the trait ──────────────────────────────────────
summary(df$Convex.Area.mm2)
hist(df$Convex.Area.mm2, breaks = 15, main = "Convex Area distribution")

# ── 1. Fit full nested model ──────────────────────────────────
convex_full <- glmmTMB(
  (Convex.Area.mm2) ~ Genotype + (1 | Batch/Plate/Seed),
  data = df, family = gaussian()
)
VarCorr(convex_full)
check_singularity(convex_full) 

# get error:   Model convergence problem; non-positive-definite Hessian matrix. See vignette('troubleshooting')
# will simplify model as std. dev. of seed:plate:batch is incredibly small (1.2363e-95)

# ── 3. Simplified model (only if full model fails) ────────────
convex_tmb <- glmmTMB(
  (Convex.Area.mm2) ~ Genotype + (1 | Batch/Plate),
  data = df, family = gaussian()
)

VarCorr(convex_tmb)
summary(convex_tmb)
check_singularity(convex_tmb) 

# plate:batch has very small effect and get NaNs inc conditional model so will drop plate effect

# Plate interaction is incredibly small so will drop 
# ── 3. Simplified model (only if simplified model fails) ────────────
convex_tmb <- glmmTMB(
  (Convex.Area.mm2) ~ Genotype + (1 | Batch),
  data = df, family = gaussian()
)

VarCorr(convex_tmb)
check_singularity(convex_tmb) 

# ── 4. Assumption checks (on whichever model passed) ─────────
sim_res <- simulateResiduals(convex_tmb, n = 1000)
plot(sim_res)                               # QQ + residuals vs fitted
testUniformity(sim_res)                     # p > 0.05 = distribution correct
testDispersion(sim_res)                     # ratio ≈ 1.0
plotResiduals(sim_res, form = df$Genotype)  # no patterns per genotype

# residuals show that there is some snaking so will try log transforming to see if fit is better 

convex_tmb <- glmmTMB( log(Convex.Area.mm2) ~ Genotype + (1 | Batch), data = df, family = gaussian() ) 
sim_res <- simulateResiduals(convex_tmb, n = 1000)
plot(sim_res)                               # QQ + residuals vs fitted
testUniformity(sim_res)                     # p > 0.05 = distribution correct
testDispersion(sim_res)                     # ratio ≈ 1.0
plotResiduals(sim_res, form = df$Genotype)  # no patterns per genotype

# imroved fit so will keep this log transformed model 

# ── 5. Test overall Genotype effect ──────────────────────────
car::Anova(convex_tmb, type = "III")


# ── 6. Pairwise comparisons + CLD ────────────────────────────
emm <- emmeans(convex_tmb, ~ Genotype, type = "response")
pairs(emm, adjust = "sidak")

cld_df <- as.data.frame(cld(emm, adjust = "sidak",
                             Letters = letters, alpha = 0.05)) %>%
  mutate(.group = trimws(.group))

n_df <- df %>%
  count(Genotype) %>%
  mutate(label = paste0("n=", n))


# ── 7. Plot 1 — Boxplot ───────────────────────────────────────

y_ceil <- max(df$Convex.Area.mm2) * 1.2
y_label <- y_ceil * 0.87

p1 <- ggplot(df, aes(x = Genotype, y = Convex.Area.mm2, fill = Genotype)) +

  geom_boxplot(
    alpha = 0.7, outlier.shape = NA,
    width = 0.5, colour = "black", linewidth = 0.5
  ) +

  geom_jitter(width = 0.15, size = 1.8, alpha = 0.55, colour = "black") +

  geom_text(
    data = cld_df, aes(x = Genotype, label = .group),
    y = y_label, inherit.aes = FALSE,
    size = 5, fontface = "bold", colour = "grey20"
  ) +

  geom_text(
    data = n_df, aes(x = Genotype, label = label),
    y = -Inf, vjust = 4.2, inherit.aes = FALSE,
    size = 3.5, colour = "grey40"
  ) +

  scale_fill_brewer(palette = "Set2") +

  labs(
    x       = "Genotype",
    y       = expression(Convex~area~(mm^2)),
    caption = "Shared letters indicate no significant difference (sidak HSD, p > 0.05; glmmTMB, Gaussian log)"
  ) +

  theme_classic(base_size = 14) +
  theme(
    legend.position = "none",
    axis.line       = element_line(colour = "black", linewidth = 0.6),
    axis.ticks      = element_line(colour = "black", linewidth = 0.5),
    axis.title      = element_text(size = 14, face = "bold"),
    axis.text       = element_text(size = 14, colour = "black"),
    axis.text.x     = element_text(face = "italic", margin = margin(t = 3)),
    plot.caption    = element_text(size = 9, colour = "grey40", hjust = 0),
    plot.margin     = margin(t = 10, r = 15, b = 30, l = 10)
  )

p1 + coord_cartesian(clip = "off")
```

### Network Area

```{r, fig.width=16, fig.height=12, dpi=300}
# ============================================================
#  Network Area Analysis
# ============================================================

# ── 1. Inspect the trait ──────────────────────────────────────
summary(df$Network.Area.mm2)
hist(df$Network.Area.mm2, breaks = 15, main = "Network Area distribution")


# ── 2. Fit full nested model first ───────────────────────────
network_full <- glmmTMB(
  (Network.Area.mm2) ~ Genotype + (1 | Batch/Plate/Seed),
  data = df, family = gaussian()
)

VarCorr(network_full)
check_singularity(network_full)
summary(network_full)

# Standard deviation is incredibly small for plate:batch so getting the Hessian matrix error so will use a model which drops this interaction


# ── 3. Simplified model (use if full model fails) ─────────────
network_tmb <- glmmTMB(
  Network.Area.mm2 ~ Genotype + (1 | Batch) + (1 | Seed:Plate:Batch),
  data = df, family = gaussian()
)

VarCorr(network_tmb)
check_singularity(network_tmb)
summary(network_tmb)   # confirm real SEs and AIC


# ── 4. Assumption checks ──────────────────────────────────────
sim_res <- simulateResiduals(network_tmb, n = 1000)
plot(sim_res)
testUniformity(sim_res)
testDispersion(sim_res)
plotResiduals(sim_res, form = df$Genotype)

# residuals show poor fit (significant deviation from outlier test) so will log transform the data

# ── 3. Log transformed model (use if full model fails) ─────────────
network_tmb <- glmmTMB(
  log(Network.Area.mm2) ~ Genotype + (1 | Batch) + (1 | Seed:Plate:Batch),
  data = df, family = gaussian()
)

VarCorr(network_tmb)
check_singularity(network_tmb)
summary(network_tmb)   # confirm real SEs and AIC


# ── 4. Assumption checks ──────────────────────────────────────
sim_res <- simulateResiduals(network_tmb, n = 1000)
plot(sim_res)
testUniformity(sim_res)
testDispersion(sim_res)
plotResiduals(sim_res, form = df$Genotype)


# ── 5. Test overall Genotype effect ──────────────────────────
car::Anova(network_tmb, type = "III")


# ── 6. Pairwise comparisons + CLD ────────────────────────────
emm <- emmeans(network_tmb, ~ Genotype, type = "response")
pairs(emm, adjust = "sidak")

cld_df <- as.data.frame(cld(emm, adjust = "sidak",
                             Letters = letters, alpha = 0.05)) %>%
  mutate(.group = trimws(.group))

n_df <- df %>%
  count(Genotype) %>%
  mutate(label = paste0("n=", n))


# ── 7. Plot 1 — Boxplot ───────────────────────────────────────

y_ceil <- max(df$Network.Area.mm2) * 1.22
y_label <- y_ceil * 0.8555

p1 <- ggplot(df, aes(x = Genotype, y = Network.Area.mm2, fill = Genotype)) +

  geom_boxplot(
    alpha = 0.7, outlier.shape = NA,
    width = 0.5, colour = "black", linewidth = 0.5
  ) +

  geom_jitter(width = 0.15, size = 1.8, alpha = 0.55, colour = "black") +

  geom_text(
    data = cld_df, aes(x = Genotype, label = .group),
    y = y_label, inherit.aes = FALSE,
    size = 5, fontface = "bold", colour = "grey20"
  ) +

  geom_text(
    data = n_df, aes(x = Genotype, label = label),
    y = -Inf, vjust = 4.2, inherit.aes = FALSE,
    size = 3.5, colour = "grey40"
  ) +

  scale_fill_brewer(palette = "Set2") +
 
  labs(
    x       = "Genotype",
    y       = expression(Network~area~(mm^2)),
    caption = "Shared letters indicate no significant difference (sidak HSD, p > 0.05; glmmTMB, Gaussian log)"
  ) +

  theme_classic(base_size = 14) +
  theme(
    legend.position = "none",
    axis.line       = element_line(colour = "black", linewidth = 0.6),
    axis.ticks      = element_line(colour = "black", linewidth = 0.5),
    axis.title      = element_text(size = 14, face = "bold"),
    axis.text       = element_text(size = 14, colour = "black"),
    axis.text.x     = element_text(face = "italic", margin = margin(t = 3)),
    plot.caption    = element_text(size = 9, colour = "grey40", hjust = 0),
    plot.margin     = margin(t = 10, r = 15, b = 30, l = 10)
  )

p1 + coord_cartesian(clip = "off")
```

### Solidity

```{r, fig.width=16, fig.height=12, dpi=300}
# ============================================================
#  Solidity Analysis
#  Solidity is a proportion (0–1) — Beta regression appropriate
# ============================================================

# ── 1. Inspect distribution ───────────────────────────────────
summary(df$Solidity)
hist(df$Solidity, breaks = 15, main = "Solidity distribution")
# Solidity is bounded 0–1 so Gaussian is inappropriate by definition
# Beta family is the correct choice for proportions


# ── 2. Check for exact 0s or 1s (Beta cannot handle these) ───
sum(df$Solidity == 0)
sum(df$Solidity == 1)
# If any exist, apply small correction:
# df$Solidity <- (df$Solidity * (nrow(df) - 1) + 0.5) / nrow(df)


# ── 3. Fit full nested model ──────────────────────────────────
solidity_tmb <- glmmTMB(
  Solidity ~ Genotype + (1 | Batch/Plate/Seed),
  data = df, family = beta_family()
)

VarCorr(solidity_tmb)
check_singularity(solidity_tmb)
summary(solidity_tmb)

# singularity is true so will reduce the model 

# ── 3. Fit reduced model ──────────────────────────────────
solidity_tmb <- glmmTMB(
  Solidity ~ Genotype + (1 | Batch/Plate),
  data = df, family = beta_family()
)

VarCorr(solidity_tmb)
check_singularity(solidity_tmb)
summary(solidity_tmb)

# singularity is false so happy to continue on to assumption checks

# ── 5. Assumption checks ──────────────────────────────────────
sim_res <- simulateResiduals(solidity_tmb, n = 1000)
plot(sim_res)                               # QQ + residuals vs fitted
testUniformity(sim_res)                     # p > 0.05 = Beta appropriate
testDispersion(sim_res)                     # ratio ≈ 1.0
plotResiduals(sim_res, form = df$Genotype)  # no patterns per genotype

# assumptions all look good so happy to continue 

# ── 6. Test overall Genotype effect ──────────────────────────
car::Anova(solidity_tmb, type = "III")


# ── 7. Pairwise comparisons + CLD ────────────────────────────
# type = "response" back-transforms from logit link to 0-1 scale
emm <- emmeans(solidity_tmb, ~ Genotype, type = "response")
pairs(emm, adjust = "sidak")

cld_df <- as.data.frame(cld(emm, adjust = "sidak",
                             Letters = letters, alpha = 0.05)) %>%
  mutate(.group = trimws(.group))

n_df <- df %>%
  count(Genotype) %>%
  mutate(label = paste0("n=", n))


# ── 8. Plot 1 — Boxplot ───────────────────────────────────────

# ── Solidity plot — y-axis zoomed to data range ───────────────

y_ceil <- max(df$Solidity) * 1.22
y_label <- y_ceil * 0.95

p1 <- ggplot(df, aes(x = Genotype, y = Solidity, fill = Genotype)) +

  geom_boxplot(
    alpha         = 0.7,
    outlier.shape = NA,
    width         = 0.5,
    colour        = "black",
    linewidth     = 0.5
  ) +

  geom_jitter(
    width  = 0.15,
    size   = 1.8,
    alpha  = 0.55,
    colour = "black"
  ) +

  geom_text(
    data        = cld_df,
    aes(x = Genotype, label = .group),
    y           = y_label,
    inherit.aes = FALSE,
    size        = 5,
    fontface    = "bold",
    colour      = "grey20"
  ) +

  geom_text(
    data        = n_df,
    aes(x = Genotype, label = label),
    y           = -Inf,
    vjust       = 4.2,
    inherit.aes = FALSE,
    size        = 3.5,
    colour      = "grey40"
  ) +

  scale_fill_brewer(palette = "Set2") +

  labs(
    x       = "Genotype",
    y       = "Solidity (proportion)",
    caption = "Shared letters indicate no significant difference (sidak HSD, p > 0.05; glmmTMB, Beta)"
  ) +

  theme_classic(base_size = 14) +
  theme(
    legend.position = "none",
    axis.line       = element_line(colour = "black", linewidth = 0.6),
    axis.ticks      = element_line(colour = "black", linewidth = 0.5),
    axis.title      = element_text(size = 14, face = "bold"),
    axis.text       = element_text(size = 14, colour = "black"),
    axis.text.x     = element_text(face = "italic", margin = margin(t = 3)),
    plot.caption    = element_text(size = 9, colour = "grey40", hjust = 0),
    plot.margin     = margin(t = 10, r = 15, b = 30, l = 10)
  )

p1 + coord_cartesian(
  ylim = c(0, max(df$Solidity) * 1.22),
  clip = "off"
)

```

### Width to Depth Ratio

```{r, fig.width=16, fig.height=12, dpi=300}
# ============================================================
#  Width-to-Depth Ratio Analysis
#  Ratio data — positive continuous, likely right-skewed
#  Start raw, log-transform only if DHARMa flags poor fit
# ============================================================

# ── 1. Inspect distribution ───────────────────────────────────
summary(df$`Width.to.Depth.Ratio`)
class(df$`Width.to.Depth.Ratio`)
hist(df$`Width.to.Depth.Ratio`, breaks = 15, main = "Width-to-Depth Ratio distribution")


# ── 2. Fit full nested model on raw scale ────────────────────
wdr_full <- glmmTMB(
  `Width.to.Depth.Ratio` ~ Genotype + (1 | Batch/Plate/Seed),
  data = df, family = gaussian()
)

VarCorr(wdr_full)
check_singularity(wdr_full)
summary(wdr_full)

# check assumptions 
sim_res <- simulateResiduals(wdr_full, n = 1000)
plot(sim_res)                               # QQ + residuals vs fitted
testUniformity(sim_res)                     # p > 0.05 = raw Gaussian fine
testDispersion(sim_res)                     # ratio ≈ 1.0
plotResiduals(sim_res, form = df$Genotype)  # no patterns per genotype

# singularity = true and poor model fit so trying reduced model 

# try fitting reduced model
wdr_tmb <- glmmTMB(
  `Width.to.Depth.Ratio` ~ Genotype + (1 | Batch/Plate),
  data = df, family = gaussian()
)

VarCorr(wdr_tmb)
check_singularity(wdr_tmb)
summary(wdr_tmb)

# this model has singularity = true as the random effects are  close to zero but model fit was good and improtant to include some random effects so will continue with this model.

# ── 4. Assumption checks ─────
sim_res <- simulateResiduals(wdr_tmb, n = 1000)
plot(sim_res)                               # QQ + residuals vs fitted
testUniformity(sim_res)                     # p > 0.05 = raw Gaussian fine
testDispersion(sim_res)                     # ratio ≈ 1.0
plotResiduals(sim_res, form = df$Genotype)  # no patterns per genotype



# ── 5. Test overall Genotype effect ──────────────────────────
car::Anova(wdr_tmb, type = "III")


# ── Pairwise comparisons + CLD ────────────────────────────────
emm <- emmeans(wdr_tmb, ~ Genotype, type = "response")
pairs(emm, adjust = "sidak")

cld_df <- as.data.frame(cld(emm, adjust = "sidak",
                             Letters = letters, alpha = 0.05)) %>%
  mutate(.group = trimws(.group))

n_df <- df %>%
  count(Genotype) %>%
  mutate(label = paste0("n=", n))

# ── Plot ──────────────────────────────────────────────────────

y_ceil <- max(df$Width.to.Depth.Ratio) * 1.22
y_label <- y_ceil * 0.95

p1 <- ggplot(df, aes(x = Genotype, y = Width.to.Depth.Ratio, fill = Genotype)) +

  geom_boxplot(
    alpha         = 0.7,
    outlier.shape = NA,
    width         = 0.5,
    colour        = "black",
    linewidth     = 0.5
  ) +

  geom_jitter(
    width  = 0.15,
    size   = 1.8,
    alpha  = 0.55,
    colour = "black"
  ) +

  geom_text(
    data        = cld_df,
    aes(x = Genotype, label = .group),
    y           = y_label,
    inherit.aes = FALSE,
    size        = 5,
    fontface    = "bold",
    colour      = "grey20"
  ) +

  geom_text(
    data        = n_df,
    aes(x = Genotype, label = label),
    y           = -Inf,
    vjust       = 4.2,
    inherit.aes = FALSE,
    size        = 3.5,
    colour      = "grey40"
  ) +

  scale_fill_brewer(palette = "Set2") +

  labs(
    x       = "Genotype",
    y       = "Width-to-depth ratio",
    caption = "Shared letters indicate no significant difference (sidak HSD, p > 0.05; glmmTMB, Gaussian)"
  ) +

  theme_classic(base_size = 14) +
  theme(
    legend.position = "none",
    axis.line       = element_line(colour = "black", linewidth = 0.6),
    axis.ticks      = element_line(colour = "black", linewidth = 0.5),
    axis.title      = element_text(size = 14, face = "bold"),
    axis.text       = element_text(size = 14, colour = "black"),
    axis.text.x     = element_text(face = "italic", margin = margin(t = 3)),
    plot.caption    = element_text(size = 9, colour = "grey40", hjust = 0),
    plot.margin     = margin(t = 10, r = 15, b = 30, l = 10)
  )

p1 + coord_cartesian(
  ylim = c(0, max(df$Width.to.Depth.Ratio) * 1.22),
  clip = "off"
)

# this is the same result as the model with no random effects included so happy to continue with this result.
```

### Steep Angle Frequency

```{r, fig.width=16, fig.height=12, dpi=300}
# ============================================================
#  Steep Angle Frequency Analysis
#  Frequency/proportion bounded 0-1 — Beta regression
#  Check for exact 0s or 1s before fitting
# ============================================================

# Per-genotype Var/Mean — Poisson assumes Var = Mean exactly.
# Ratios >> 1 within genotypes suggest NB is needed.
df %>%
  group_by(Genotype) %>%
  summarise(Mean     = round(mean(Steep.Angle.Frequency), 2),
            Variance = round(var(Steep.Angle.Frequency),  2),
            Var_Mean = round(var(Steep.Angle.Frequency) /
                               mean(Steep.Angle.Frequency), 2))

# ── 1. Inspect distribution ───────────────────────────────────
summary(df$Steep.Angle.Frequency)
hist(df$Steep.Angle.Frequency, breaks = 15, main = "Steep Angle Frequency distribution")

# Check for exact 0s or 1s — Beta cannot handle boundary values
sum(df$Steep.Angle.Frequency == 0)
sum(df$Steep.Angle.Frequency == 1)

# If any exist, apply small boundary correction:
# df$Steep.Angle.Frequency <- (df$Steep.Angle.Frequency * (nrow(df) - 1) + 0.5) / nrow(df)


# ── 2. Fit full nested model ──────────────────────────────────
steep_full <- glmmTMB(
  Steep.Angle.Frequency ~ Genotype + (1 | Batch/Plate/Seed),
  data = df, family = beta_family()
)

VarCorr(steep_full)
check_singularity(steep_full)
summary(steep_full)

# assumptions

sim_res <- simulateResiduals(steep_full, n = 1000)
plot(sim_res)
testUniformity(sim_res)                     # p > 0.05 = Beta appropriate
testDispersion(sim_res)                     # ratio ≈ 1.0
plotResiduals(sim_res, form = df$Genotype)  # no patterns per genotype


# try reducing the model
steep_tmb <- glmmTMB(
  Steep.Angle.Frequency ~ Genotype + (1 | Batch/Plate),
  data = df, family = beta_family()
)

VarCorr(steep_tmb)
check_singularity(steep_tmb)
summary(steep_tmb)

steep_tmb <- glmmTMB(
  Steep.Angle.Frequency ~ Genotype + (1 | Batch),
  data = df, family = beta_family()
)

VarCorr(steep_tmb)
check_singularity(steep_tmb)
summary(steep_tmb)

steep_tmb <- glmmTMB(
  Steep.Angle.Frequency ~ Genotype,
  data = df, family = beta_family()
)

summary(steep_tmb)

# reduced model does not have homogeneity of variance so full model is probably better. I will try both reduced and full model to see if same result is drawn.


# Assumption checks

sim_res <- simulateResiduals(steep_tmb, n = 1000)
plot(sim_res)
testUniformity(sim_res)                     # p > 0.05 = Beta appropriate
testDispersion(sim_res)                     # ratio ≈ 1.0
plotResiduals(sim_res, form = df$Genotype)  # no patterns per genotype


# ── 5. Test overall Genotype effect ──────────────────────────
car::Anova(steep_tmb, type = "III")


# ── 6. Pairwise comparisons + CLD ────────────────────────────
emm <- emmeans(steep_tmb, ~ Genotype, type = "response")
pairs(emm, adjust = "sidak")

cld_df <- as.data.frame(cld(emm, adjust = "sidak",
                             Letters = letters, alpha = 0.05)) %>%
  mutate(.group = trimws(.group))

class(df$Genotype)   # should return "factor"
length(df$Genotype)  # should return 71


# Effect size

# Cohen's d = difference in means / pooled SD
# Interpretation: 0.2 = small, 0.5 = medium, 0.8 = large

emm_steep <- emmeans(steep_tmb, ~ Genotype)

eff_steep <- eff_size(
  emm_steep,
  sigma = sigma(steep_tmb),   # residual SD from model
  edf   = df.residual(steep_tmb)
)
print(eff_steep)

# ── 7. Plot 1 — Boxplot ───────────────────────────────────────

y_ceil <- max(df$Steep.Angle.Frequency) * 1.22
y_label <- y_ceil * 0.95

p1 <- ggplot(df, aes(x = Genotype, y = Steep.Angle.Frequency, fill = Genotype)) +

  geom_boxplot(
    alpha         = 0.7,
    outlier.shape = NA,
    width         = 0.5,
    colour        = "black",
    linewidth     = 0.5
  ) +

  geom_jitter(
    width  = 0.15,
    size   = 1.8,
    alpha  = 0.55,
    colour = "black"
  ) +

  geom_text(
    data        = cld_df,
    aes(x = Genotype, label = .group),
    y           = y_label,
    inherit.aes = FALSE,
    size        = 5,
    fontface    = "bold",
    colour      = "grey20"
  ) +

  geom_text(
    data        = n_df,
    aes(x = Genotype, label = label),
    y           = -Inf,
    vjust       = 4.2,
    inherit.aes = FALSE,
    size        = 3.5,
    colour      = "grey40"
  ) +

  scale_fill_brewer(palette = "Set2") +

  labs(
    x       = "Genotype",
    y       = "Steep angle frequency",
    caption = "Shared letters indicate no significant difference (sidak HSD, p > 0.05; glmmTMB, Beta)"
  ) +

  theme_classic(base_size = 14) +
  theme(
    legend.position = "none",
    axis.line       = element_line(colour = "black", linewidth = 0.6),
    axis.ticks      = element_line(colour = "black", linewidth = 0.5),
    axis.title      = element_text(size = 14, face = "bold"),
    axis.text       = element_text(size = 14, colour = "black"),
    axis.text.x     = element_text(face = "italic", margin = margin(t = 3)),
    plot.caption    = element_text(size = 9, colour = "grey40", hjust = 0),
    plot.margin     = margin(t = 10, r = 15, b = 30, l = 10)
  )

p1 + coord_cartesian(
  ylim = c(0, max(df$Steep.Angle.Frequency) * 1.22),
  clip = "off"
)

# full model

# ── 6. Pairwise comparisons + CLD ────────────────────────────
emm <- emmeans(steep_full, ~ Genotype, type = "response")
pairs(emm, adjust = "sidak")

cld_df <- as.data.frame(cld(emm, adjust = "sidak",
                             Letters = letters, alpha = 0.05)) %>%
  mutate(.group = trimws(.group))

# Effect size

# Cohen's d = difference in means / pooled SD
# Interpretation: 0.2 = small, 0.5 = medium, 0.8 = large

emm_steep <- emmeans(steep_full, ~ Genotype)

eff_steep <- eff_size(
  emm_steep,
  sigma = sigma(steep_full),   # residual SD from model
  edf   = df.residual(steep_full)
) 
print(eff_steep)

# ── 7. Plot 1 — Boxplot ───────────────────────────────────────

y_ceil <- max(df$Steep.Angle.Frequency) * 1.22
y_label <- y_ceil * 0.95

p1 <- ggplot(df, aes(x = Genotype, y = Steep.Angle.Frequency, fill = Genotype)) +

  geom_boxplot(
    alpha         = 0.7,
    outlier.shape = NA,
    width         = 0.5,
    colour        = "black",
    linewidth     = 0.5
  ) +

  geom_jitter(
    width  = 0.15,
    size   = 1.8,
    alpha  = 0.55,
    colour = "black"
  ) +

  geom_text(
    data        = cld_df,
    aes(x = Genotype, label = .group),
    y           = y_label,
    inherit.aes = FALSE,
    size        = 8,
    fontface    = "bold",
    colour      = "grey20"
  ) +

  geom_text(
    data        = n_df,
    aes(x = Genotype, label = label),
    y           = -Inf,
    vjust       = 4.2,
    inherit.aes = FALSE,
    size        = 6,
    colour      = "grey40"
  ) +

  scale_fill_brewer(palette = "Set2") +

  labs(
    x       = "Genotype",
    y       = "Steep angle frequency",
    caption = "Shared letters indicate no significant difference (sidak HSD, p > 0.05; glmmTMB, Beta)"
  ) +

  theme_classic(base_size = 14) +
  theme(
    legend.position = "none",
    panel.grid      = element_blank(),
    axis.line       = element_line(colour = "black", linewidth = 0.6),
    axis.ticks      = element_line(colour = "black", linewidth = 0.5),
    axis.title      = element_text(size = 26, face = "bold"),
    axis.title.x    = element_text(size = 26, face = "bold", margin = margin(t = 24)),
    axis.text       = element_text(size = 22, colour = "black"),
    axis.text.x     = element_text(face = "italic", margin = margin(t = 15)),
    plot.caption    = element_text(size = 12, colour = "grey40", hjust = 0, margin = margin(t = 10)),
    plot.margin     = margin(t = 25, r = 25, b = 25, l = 25)
  )


p1 + coord_cartesian(
  ylim = c(0, max(df$Steep.Angle.Frequency) * 1.22),
  clip = "off"
)

# difference here is where Dabbi groups - will go with full model for the figure as the model fit was better but good to see largely the same significant differences. 
```

### Depth

```{r, fig.width=16, fig.height=12, dpi=300}
# ============================================================
#  Depth Analysis
#  Positive continuous — start raw Gaussian, log if needed
# ============================================================

# ── 1. Inspect distribution ───────────────────────────────────
summary(df$Depth.mm)
hist(df$Depth.mm, breaks = 15, main = "Depth distribution")


# ── 2. Fit full nested model ──────────────────────────────────
depth_full <- glmmTMB(
  Depth.mm ~ Genotype + (1 | Batch/Plate/Seed),
  data = df, family = gaussian()
)

VarCorr(depth_full)
check_singularity(depth_full)
summary(depth_full)

# fit looks good so happy to continue with this but will do some assumption checks 

# ── 4. Assumption checks ─────
sim_res <- simulateResiduals(depth_full, n = 1000)
plot(sim_res)
testUniformity(sim_res)                     # p > 0.05 = raw Gaussian fine
testDispersion(sim_res)                     # ratio ≈ 1.0
plotResiduals(sim_res, form = df$Genotype)


# ── 5. Test overall Genotype effect ──────────────────────────
car::Anova(depth_full, type = "III")


# ── 6. Pairwise comparisons + CLD ────────────────────────────
emm <- emmeans(depth_full, ~ Genotype)
pairs(emm, adjust = "sidak")

cld_df <- as.data.frame(cld(emm, adjust = "sidak",
                             Letters = letters, alpha = 0.05)) %>%
  mutate(.group = trimws(.group))

n_df <- df %>%
  count(Genotype) %>%
  mutate(label = paste0("n=", n))


# ── 7. Plot 1 — Boxplot ───────────────────────────────────────

y_ceil <- max(df$Depth.mm) * 1.22
y_label <- y_ceil * 0.95

p1 <- ggplot(df, aes(x = Genotype, y = Depth.mm, fill = Genotype)) +

  geom_boxplot(
    alpha         = 0.7,
    outlier.shape = NA,
    width         = 0.5,
    colour        = "black",
    linewidth     = 0.5
  ) +

  geom_jitter(
    width  = 0.15,
    size   = 1.8,
    alpha  = 0.55,
    colour = "black"
  ) +

  geom_text(
    data        = cld_df,
    aes(x = Genotype, label = .group),
    y           = y_label,
    inherit.aes = FALSE,
    size        = 5,
    fontface    = "bold",
    colour      = "grey20"
  ) +

  geom_text(
    data        = n_df,
    aes(x = Genotype, label = label),
    y           = -Inf,
    vjust       = 4.2,
    inherit.aes = FALSE,
    size        = 3.5,
    colour      = "grey40"
  ) +

  scale_fill_brewer(palette = "Set2") +

  labs(
    x       = "Genotype",
    y       = "Depth (mm)",
    caption = "Shared letters indicate no significant difference (sidak HSD, p > 0.05; glmmTMB, Gaussian)"
  ) +

  theme_classic(base_size = 14) +
  theme(
    legend.position = "none",
    axis.line       = element_line(colour = "black", linewidth = 0.6),
    axis.ticks      = element_line(colour = "black", linewidth = 0.5),
    axis.title      = element_text(size = 14, face = "bold"),
    axis.text       = element_text(size = 14, colour = "black"),
    axis.text.x     = element_text(face = "italic", margin = margin(t = 3)),
    plot.caption    = element_text(size = 9, colour = "grey40", hjust = 0),
    plot.margin     = margin(t = 10, r = 15, b = 30, l = 10)
  )

p1 + coord_cartesian(
  ylim = c(0, max(df$Depth.mm) * 1.22),
  clip = "off"
)

```

### Surface area

```{r, fig.width=16, fig.height=12, dpi=300}
# ============================================================
#  Depth Analysis
#  Positive continuous — start raw Gaussian, log if needed
# ============================================================

# ── 1. Inspect distribution ───────────────────────────────────
summary(df$Surface.Area.mm2)
hist(df$Surface.Area.mm2, breaks = 15, main = "Surface Area")


# ── 2. Fit full nested model ──────────────────────────────────
SA_full <- glmmTMB(
  Surface.Area.mm2 ~ Genotype + (1 | Batch/Plate/Seed),
  data = df, family = gaussian()
)

VarCorr(SA_full)
check_singularity(SA_full)
summary(SA_full)

# Get Hessian matrix error because plate:batch interaction is very small so will drop this interaction
SA_tmb <- glmmTMB(
  Surface.Area.mm2 ~ Genotype + (1 | Batch) + (1 | Batch:Plate:Seed),
  data = df, family = gaussian()
)

VarCorr(SA_tmb)
check_singularity(SA_tmb)
summary(SA_tmb)

# Still getting error so dropping the batch:plate:seed interaction
SA_tmb <- glmmTMB(
  Surface.Area.mm2 ~ Genotype + (1 | Batch),
  data = df, family = gaussian()
)

VarCorr(SA_tmb)
check_singularity(SA_tmb)
summary(SA_tmb)


sim_res <- simulateResiduals(SA_tmb, n = 1000)
plot(sim_res)
testUniformity(sim_res)                     # p > 0.05 = raw Gaussian fine
testDispersion(sim_res)                     # ratio ≈ 1.0
plotResiduals(sim_res, form = df$Genotype)

# fit is pretty good so happy to continue 

# ── 5. Test overall Genotype effect ──────────────────────────
car::Anova(SA_tmb, type = "III")

# ── 6. Pairwise comparisons + CLD ────────────────────────────
emm <- emmeans(SA_tmb, ~ Genotype)
pairs(emm, adjust = "sidak")

cld_df <- as.data.frame(cld(emm, adjust = "sidak",
                             Letters = letters, alpha = 0.05)) %>%
  mutate(.group = trimws(.group))

n_df <- df %>%
  count(Genotype) %>%
  mutate(label = paste0("n=", n))


# ── 7. Plot 1 — Boxplot ───────────────────────────────────────

y_ceil <- max(df$Surface.Area.mm2) * 1.22
y_label <- y_ceil * 0.95

p1 <- ggplot(df, aes(x = Genotype, y = Surface.Area.mm2, fill = Genotype)) +

  geom_boxplot(
    alpha         = 0.7,
    outlier.shape = NA,
    width         = 0.5,
    colour        = "black",
    linewidth     = 0.5
  ) +

  geom_jitter(
    width  = 0.15,
    size   = 1.8,
    alpha  = 0.55,
    colour = "black"
  ) +

  geom_text(
    data        = cld_df,
    aes(x = Genotype, label = .group),
    y           = y_label,
    inherit.aes = FALSE,
    size        = 5,
    fontface    = "bold",
    colour      = "grey20"
  ) +

  geom_text(
    data        = n_df,
    aes(x = Genotype, label = label),
    y           = -Inf,
    vjust       = 4.2,
    inherit.aes = FALSE,
    size        = 3.5,
    colour      = "grey40"
  ) +

  scale_fill_brewer(palette = "Set2") +

  labs(
    x       = "Genotype",
    y       = "Surface Area (mm)",
    caption = "Shared letters indicate no significant difference (sidak HSD, p > 0.05; glmmTMB, Gaussian)"
  ) +

  theme_classic(base_size = 14) +
  theme(
    legend.position = "none",
    axis.line       = element_line(colour = "black", linewidth = 0.6),
    axis.ticks      = element_line(colour = "black", linewidth = 0.5),
    axis.title      = element_text(size = 14, face = "bold"),
    axis.text       = element_text(size = 14, colour = "black"),
    axis.text.x     = element_text(face = "italic", margin = margin(t = 3)),
    plot.caption    = element_text(size = 9, colour = "grey40", hjust = 0),
    plot.margin     = margin(t = 10, r = 15, b = 30, l = 10)
  )

p1 + coord_cartesian(
  ylim = c(0, max(df$Surface.Area.mm2) * 1.22),
  clip = "off"
)

```

### Volume

```{r, fig.width=16, fig.height=12, dpi=300}
# ============================================================
# Volume Analysis  
# Positive continuous — start raw Gaussian, log if needed
# ============================================================

# Per-genotype Var/Mean — Poisson assumes Var = Mean exactly.
# Ratios >> 1 within genotypes suggest NB is needed.
df %>%
  group_by(Genotype) %>%
  summarise(Mean     = round(mean(Volume.mm3), 2),
            Variance = round(var(Volume.mm3),  2),
            Var_Mean = round(var(Volume.mm3) /
                               mean(Volume.mm3), 2))

# ── 1. Inspect distribution ───────────────────────────────────
summary(df$Volume.mm3)
hist(df$Volume.mm3, breaks = 15, main = "Volume (mm³)")

# ── 2. Fit models ────────────────────────────────────────
Vol_raw <- glmmTMB(
  Volume.mm3 ~ Genotype + (1 | Batch/Plate/Seed),
  data = df, 
  family = gaussian()
)

VarCorr(Vol_raw)
check_singularity(Vol_raw)
summary(Vol_raw)

# ── 4. Diagnostics: raw model ────────────────────────────────
sim_res_raw <- simulateResiduals(Vol_raw, n = 1000)
plot(sim_res_raw)
testUniformity(sim_res_raw)    # p > 0.05 = good
testDispersion(sim_res_raw)    # ratio ≈ 1.0
plotResiduals(sim_res_raw, form = df$Genotype)

# some snaking so will try log transforming to see if there is an improved fit
Vol_raw <- glmmTMB(
  log(Volume.mm3) ~ Genotype + (1 | Batch/Plate/Seed),
  data = df, 
  family = gaussian()
)

VarCorr(Vol_raw)
summary(Vol_raw)
check_singularity(Vol_raw)

sim_res_raw <- simulateResiduals(Vol_raw, n = 1000)
plot(sim_res_raw)
testUniformity(sim_res_raw)    # p > 0.05 = good
testDispersion(sim_res_raw)    # ratio ≈ 1.0
plotResiduals(sim_res_raw, form = df$Genotype)

# overall fit with log transformation ooks better but there is within-group deviations from uniformity and non singular so will keep the non-log transformed model 

Vol_tmb <- glmmTMB(
  Volume.mm3 ~ Genotype + (1 | Batch/Plate/Seed),
  data = df, 
  family = gaussian()
)

# ── 5. Diagnostics: log model ────────────────────────────────
sim_res_vol <- simulateResiduals(Vol_tmb, n = 1000)
plot(sim_res_vol)
testUniformity(sim_res_vol)    # p > 0.05 = good
testDispersion(sim_res_vol)    # ratio ≈ 1.0
plotResiduals(sim_res_vol, form = df$Genotype)


car::Anova(Vol_tmb, type = "III")

# Pairwise (type = "response" ONLY if using log model)
emm <- emmeans(Vol_tmb, ~ Genotype, 
               type = "response")
pairs(emm, adjust = "sidak")

# Cohen's d = difference in means / pooled SD
# Interpretation: 0.2 = small, 0.5 = medium, 0.8 = large

eff_vol <- eff_size(
  emm,
  sigma = sigma(Vol_tmb),   # residual SD from model
  edf   = df.residual(Vol_tmb)
)
print(eff_vol)

# ── 8. CLD and sample sizes ──────────────────────────────────
cld_df <- as.data.frame(cld(emm, adjust = "sidak",
                             Letters = letters, alpha = 0.05)) %>%
  mutate(.group = trimws(.group))

n_df <- df %>%
  count(Genotype) %>%
  mutate(label = paste0("n=", n))

# ── 9. Plot — Boxplot ────────────────────────────────────────

y_ceil <- max(df$Volume.mm3) * 1.22
y_label <- y_ceil * 0.95

p1 <- ggplot(df, aes(x = Genotype, y = Volume.mm3, fill = Genotype)) +
  
  geom_boxplot(
    alpha         = 0.7,
    outlier.shape = NA,
    width         = 0.5,
    colour        = "black",
    linewidth     = 0.5
  ) +
  
  geom_jitter(
    width  = 0.15,
    size   = 1.8,
    alpha  = 0.55,
    colour = "black"
  ) +
  
  geom_text(
    data        = cld_df,
    aes(x = Genotype, label = .group),
    y           = y_label,
    inherit.aes = FALSE,
    size        = 8,
    fontface    = "bold",
    colour      = "grey20"
  ) +
  
  geom_text(
    data        = n_df,
    aes(x = Genotype, label = label),
    y           = -Inf,
    vjust       = 4.2,
    inherit.aes = FALSE,
    size        = 6,
    colour      = "grey40"
  ) +
  
  scale_fill_brewer(palette = "Set2") +

  labs(
    x       = "Genotype",
    y       = "Volume (mm³)",
    caption = "Shared letters indicate no significant difference (sidak HSD, p > 0.05; glmmTMB, Gaussian)"
  ) +
  
  theme_classic(base_size = 14) +
  theme(
    legend.position = "none",
    panel.grid      = element_blank(),
    axis.line       = element_line(colour = "black", linewidth = 0.6),
    axis.ticks      = element_line(colour = "black", linewidth = 0.5),
    axis.title      = element_text(size = 26, face = "bold"),
    axis.title.x    = element_text(size = 26, face = "bold", margin = margin(t = 24)),
    axis.text       = element_text(size = 22, colour = "black"),
    axis.text.x     = element_text(face = "italic", margin = margin(t = 15)),
    plot.caption    = element_text(size = 12, colour = "grey40", hjust = 0, margin = margin(t = 10)),
    plot.margin     = margin(t = 25, r = 25, b = 25, l = 25)
  )

p1 + coord_cartesian(
  ylim = c(0, max(df$Volume.mm3) * 1.22),
  clip = "off"
)

```
